---
title: "PBTA-germline demographic and clinical statistics"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2023
params:
  plot_ci: TRUE
---

This script generates demographic and clinical summary statistics for patients with and without germline pathogenic/likely pathogenic (P/LP) variants in cancer prediposition genes (CPGs)

## Setup

#### Packages and functions

Load packages and set directory paths

```{r}
library(tidyverse)
library(openxlsx)
library(ComplexHeatmap)
library(circlize)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "demo-clin-stats")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# If the input and results directories do not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

# Load function script for generating summary tables
source(file.path(root_dir, "figures", "theme.R"))
source(file.path(analysis_dir, "util", "summary_functions.R"))
source(file.path(analysis_dir, "util", "heatmap_function.R"))
```

Set file paths

```{r set paths}
cbtn_histologies_file <- file.path(root_dir, "analyses", 
                                   "collapse-tumor-histologies", "results", 
                                   "germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

all_plp_file <- file.path(data_dir, 
                      "pbta-merged-plp-variants-autogvp-abridged.tsv")

plp_sv_file <- file.path(data_dir, "pbta_germline_svs.tsv")

cpg_file <- file.path(root_dir, "analyses", "oncokb-annotation", "input", "cpg.txt")
```

Wrangle data

```{r}
hist <- read_tsv(cbtn_histologies_file) %>%
  dplyr::mutate(OS_years = OS_days/365.25,
                EFS_years = EFS_days/365.25)

all_plp <- read_tsv(all_plp_file) %>%
  dplyr::filter(Kids_First_Biospecimen_ID_normal %in% hist$Kids_First_Biospecimen_ID_normal)

plp_sv <- read_tsv(plp_sv_file) %>%
  dplyr::filter(Kids_First_Biospecimen_ID_normal %in% hist$Kids_First_Biospecimen_ID_normal)

cpgs <- read_lines(cpg_file)

tmb <- read_tsv(file.path(data_dir, "snv-mutation-tmb-coding.tsv")) %>%
  dplyr::rename("Kids_First_Biospecimen_ID_tumor" = Tumor_Sample_Barcode)
```

Merge sample metadata

```{r}
# Extract BS IDs of samples with CPG PLP variants
cpg_bs <- all_plp %>%
  dplyr::filter(gene_symbol_vep %in% cpgs) %>%
  dplyr::select(Kids_First_Biospecimen_ID_normal) %>%
  bind_rows(plp_sv %>% dplyr::select(Kids_First_Biospecimen_ID_normal)) %>%
  pull(Kids_First_Biospecimen_ID_normal)

# Append CPG PLP status and TMB data to `hist`
hist <- hist %>%
  dplyr::mutate(cpgPLP_status = case_when(
    Kids_First_Biospecimen_ID_normal %in% cpg_bs ~ "cpgPLP",
    !Kids_First_Biospecimen_ID_normal %in% cpg_bs ~ "no_cpgPLP"
  )) %>%
  left_join(tmb[,c("Kids_First_Biospecimen_ID_tumor", "tmb")], by = "Kids_First_Biospecimen_ID_tumor")

# Define OS years and EFS years
hist <- hist %>%
  dplyr::mutate(OS_years = as.numeric(OS_days)/365.25,
                EFS_years = as.numeric(EFS_days)/365.25)
```

Merge other, unreported, and unknown race and ethnicities 

```{r}
unk_race <- c("Not Reported", "Other", "Reported Unknown")
unk_ethnicity <- c("Not Available", "Not Reported", "Unavailable/Not Reported")

hist <- hist %>%
    dplyr::mutate(race = case_when(
      race %in% unk_race ~ "Other/Unknown Race",
      !race %in% unk_race ~ race
    )) %>%
    mutate(ethnicity = case_when(
      ethnicity %in% unk_ethnicity ~ "Other/Unknown Ethnicity",
      !ethnicity %in% unk_ethnicity ~ ethnicity
    )) %>%
  dplyr::mutate(plot_group = case_when(
    plot_group == "Atypical Teratoid Rhabdoid Tumor" ~ "ATRT",
    TRUE ~ plot_group
  ))
```
  
Obtain count and median statistics on demographic and clinical data for patients with and without germline CPG P/LP variants
  
```{r}
# Get count and percent of patients with and without P/LP variants
freq <- summarize_freq(hist, "cpgPLP_status")

# Look at distribution of following variables among patients with and without P/LP variants: sex, reported race, predicted ancestry, reported ethnicity, and histology group
sex <- summarize_count(hist, "cpgPLP_status", "germline_sex_estimate")

race <- summarize_count(hist, "cpgPLP_status", "race") %>%
  select(-cpgPLP_status)

ancestry <- summarize_count(hist, "cpgPLP_status", "predicted_ancestry") %>%
  select(-cpgPLP_status)

ethnicity <- summarize_count(hist, "cpgPLP_status", "ethnicity") %>%
  select(-cpgPLP_status)

group <- summarize_count(hist, "cpgPLP_status", "plot_group") %>%
  select(-cpgPLP_status)

# merge stats and add %
demo_path <- sex %>%
  bind_cols(race, ethnicity, ancestry, group) %>%
  column_to_rownames("cpgPLP_status") %>%
  t() %>%
  as.data.frame() %>%
  dplyr::mutate(perc_cpgPLP = round(cpgPLP/sum(hist$cpgPLP_status == "cpgPLP")*100, 1),
                perc_no_cpgPLP = round(no_cpgPLP/sum(hist$cpgPLP_status == "no_cpgPLP")*100, 1)) %>%
  dplyr::mutate(cpgPLP = glue::glue("{cpgPLP} ({perc_cpgPLP}%)"),
                no_cpgPLP = glue::glue("{no_cpgPLP} ({perc_no_cpgPLP}%)")) %>%
  dplyr::select(-perc_cpgPLP, -perc_no_cpgPLP) %>%
  dplyr::mutate(cpgPLP = case_when(
    grepl("NA", cpgPLP) ~ "0 (0%)",
    TRUE ~ cpgPLP
  ))
  
# Calculate median survival and tmb for each group
clin <- hist %>%
  group_by(cpgPLP_status) %>%
  summarise(median_age = round(median(as.numeric(age_at_diagnosis_days)/365.25, na.rm = T), 2),
            median_os_years = round(median(OS_years, na.rm = T), 2),
            median_efs_years = round(median(EFS_years, na.rm = T), 2),
            median_tmb = round(median(tmb, na.rm = T), 2)) %>%
  column_to_rownames("cpgPLP_status") %>%
  t() %>%
  as.data.frame() %>%
  dplyr::mutate(cpgPLP = as.character(cpgPLP),
                no_cpgPLP = as.character(no_cpgPLP))

# Merge stats and save
stats <- freq %>%
  bind_rows(demo_path, clin) %>%
  rownames_to_column("Term")

write_tsv(stats, 
          file.path(results_dir, "demo-clin-stats-all.tsv"))
```

Obtain p-values through appropriate tests for each term (fisher for count data, wilcox.test for continuous data)

```{r}
terms = c("sex", "race", "ancestry", "ethnicity",
          "group", "age", "OS_years", "EFS_years", "tmb")

sex_p <- fisher.test(table(hist$germline_sex_estimate, hist$cpgPLP_status))$p.value

race_p <- fisher.test(table(hist$race, hist$cpgPLP_status))$p.value

ancestry_p <- fisher.test(table(hist$predicted_ancestry, hist$cpgPLP_status))$p.value

ethnicity_p <- fisher.test(table(hist$ethnicity, hist$cpgPLP_status))$p.value

# compute p-value by Monte Carlo simulation here due to large table size
group_p <- fisher.test(table(hist$plot_group, hist$cpgPLP_status), simulate.p.value = TRUE)$p.value

age_p <- wilcox.test(as.numeric(hist$age_at_diagnosis_days[hist$cpgPLP_status == "cpgPLP"]),
                     as.numeric(hist$age_at_diagnosis_days[hist$cpgPLP_status == "no_cpgPLP"]))$p.value

os_p <- wilcox.test(hist$OS_years[hist$cpgPLP_status == "cpgPLP"],
          hist$OS_years[hist$cpgPLP_status == "no_cpgPLP"])$p.value

efs_p <- wilcox.test(hist$EFS_years[hist$cpgPLP_status == "cpgPLP"],
          hist$EFS_years[hist$cpgPLP_status == "no_cpgPLP"])$p.value

tmb_p <- wilcox.test(hist$tmb[hist$cpgPLP_status == "cpgPLP"],
          hist$tmb[hist$cpgPLP_status == "no_cpgPLP"])$p.value

# Merge p-values and write to output
pvals <- data.frame(term = terms,
                    pvalue = c(sex_p, race_p, ancestry_p, ethnicity_p, 
                               group_p, age_p, os_p, efs_p, tmb_p))

write_tsv(pvals, 
          file.path(results_dir, "demo-clin-pvals-all.tsv"))
```

Obtain same summary statistics for each histology group

```{r}
groups <- unique(hist$plot_group)

# remove `Other CNS embryonal tumor`, since there are no patients with P/LP variants
groups <- groups[groups != "Other CNS embryonal tumor"]

# Create empty list to store stats
stats_list <- list()

for (group in groups) {
  
  group_hist <- hist %>%
    filter(plot_group == group)
  
  freq <- summarize_freq(group_hist, "cpgPLP_status")

  sex <- summarize_count(group_hist, "cpgPLP_status", "germline_sex_estimate")
  
  race <- summarize_count(group_hist, "cpgPLP_status", "race") %>%
    select(-cpgPLP_status)
  
  ancestry <- summarize_count(group_hist, "cpgPLP_status", "predicted_ancestry") %>%
    select(-cpgPLP_status)
  
  ethnicity <- summarize_count(group_hist, "cpgPLP_status", "ethnicity") %>%
    select(-cpgPLP_status)
 
  demo_path <- sex %>%
    bind_cols(race, ethnicity, ancestry) %>%
    column_to_rownames("cpgPLP_status") %>%
    t() %>%
    as.data.frame() %>%
    dplyr::mutate(perc_cpgPLP = round(cpgPLP/sum(group_hist$cpgPLP_status == "cpgPLP")*100, 1),
                  perc_no_cpgPLP = round(no_cpgPLP/sum(group_hist$cpgPLP_status == "no_cpgPLP")*100, 1)) %>%
    dplyr::mutate(cpgPLP = glue::glue("{cpgPLP} ({perc_cpgPLP}%)"),
                  no_cpgPLP = glue::glue("{no_cpgPLP} ({perc_no_cpgPLP}%)")) %>%
    dplyr::select(-perc_cpgPLP, -perc_no_cpgPLP) %>%
    dplyr::mutate(cpgPLP = case_when(
      grepl("NA", cpgPLP) ~ "0 (0%)",
      TRUE ~ cpgPLP
    ))
  
  clin <- group_hist %>%
    group_by(cpgPLP_status) %>%
    summarise(median_age = round(median(as.numeric(age_at_diagnosis_days)/365.25, na.rm = T), 2),
              median_os_years = round(median(OS_years, na.rm = T), 2),
              median_efs_years = round(median(EFS_years, na.rm = T), 2),
              median_tmb = round(median(tmb, na.rm = T), 2)) %>%
    column_to_rownames("cpgPLP_status") %>%
    t() %>%
    as.data.frame() %>%
    dplyr::mutate(cpgPLP = as.character(cpgPLP),
                  no_cpgPLP = as.character(no_cpgPLP))
  
  stats_list[[group]] <- freq %>%
    bind_rows(demo_path, clin)

}

# write stats by group to excel file, with separate tab for each group
write.xlsx(stats_list,
           file.path(results_dir, "demo-clin-stats-by-histology.xlsx"),
           overwrite = TRUE,
           keepNA = TRUE,
           rowNames = TRUE)
```

Obtain corresponding p-values for terms by histology

```{r}
pvalue_list <- list()

terms = c("age", "sex", "race", "ancestry", "ethnicity",
          "OS_years", "EFS_years", "tmb")

for (group in groups) {
  
  group_hist <- hist %>%
    filter(plot_group == group)
  
  sex_p <- fisher.test(table(group_hist$germline_sex_estimate, group_hist$cpgPLP_status))$p.value
  
  race_p <- fisher.test(table(group_hist$race, group_hist$cpgPLP_status))$p.value
  
  ancestry_p <- fisher.test(table(group_hist$predicted_ancestry, group_hist$cpgPLP_status))$p.value
  
  ethnicity_p <- fisher.test(table(group_hist$ethnicity, group_hist$cpgPLP_status))$p.value
  
  age_p <- wilcox.test(as.numeric(group_hist$age_at_diagnosis_days[group_hist$cpgPLP_status == "cpgPLP"]),
                       as.numeric(group_hist$age_at_diagnosis_days[group_hist$cpgPLP_status == "no_cpgPLP"]))$p.value

  os_p <- wilcox.test(group_hist$OS_years[group_hist$cpgPLP_status == "cpgPLP"],
            group_hist$OS_years[group_hist$cpgPLP_status == "no_cpgPLP"])$p.value
  
  efs_p <- wilcox.test(group_hist$EFS_years[group_hist$cpgPLP_status == "cpgPLP"],
            group_hist$EFS_years[group_hist$cpgPLP_status == "no_cpgPLP"])$p.value
  
  tmb_p <- wilcox.test(group_hist$tmb[hist$cpgPLP_status == "cpgPLP"],
            group_hist$tmb[hist$cpgPLP_status == "no_cpgPLP"])$p.value
  
  pvalue_list[[group]] <- data.frame(term = terms,
                                    pvalue = c(age_p, sex_p, race_p, ancestry_p,
                                               ethnicity_p, os_p, efs_p, tmb_p))
}

write.xlsx(pvalue_list,
           file.path(results_dir, "demo-clin-pvalues-by-histology.xlsx"),
           overwrite = TRUE,
           keepNA = TRUE,
           rowNames = FALSE)
```

Plot tumor histology enrichment for P/LP carriers

```{r}
ht_group <- plot_enr(hist, "plot_group", "cpgPLP_status",
                     var1_names = sort(unique(hist$plot_group)), 
                     var2_names = unique(hist$cpgPLP_status),
                     padjust = FALSE)

pdf(file.path(plot_dir, "plot_group_plp_carrier_enr_heatmap.pdf"),
    width = 5, height = 8)

draw(ht_group)

dev.off()
```

Plot P-LP carrier enrichment for tumor molecular subtypes

```{r}
hist <- hist %>%
    mutate(mol_sub_group = case_when(
    plot_group %in% c("Low-grade glioma") & grepl("SEGA", molecular_subtype) ~ "SEGA",
    plot_group %in% c("Low-grade glioma") & grepl("BRAF V600E", molecular_subtype) ~ "LGG, BRAF V600E",
    plot_group %in% c("Low-grade glioma") & grepl("-BRAF", molecular_subtype) ~ "LGG, BRAF fusion",
    plot_group %in% c("Low-grade glioma") & grepl("wildtype", molecular_subtype) ~ "LGG, WT",
    plot_group %in% c("Low-grade glioma") ~ "LGG, Other alteration",
    plot_group %in% c("Mixed neuronal-glial tumor") & grepl("wildtype", molecular_subtype) ~ "GNG/GNT, WT",
    plot_group %in% c("Mixed neuronal-glial tumor") & grepl("BRAF V600E", molecular_subtype) ~ "GNG/GNT, BRAF V600E",
    plot_group %in% c("Mixed neuronal-glial tumor") & grepl("-BRAF", molecular_subtype) ~ "GNG/GNT, BRAF fusion",
    plot_group %in% c("Mixed neuronal-glial tumor") & !grepl("To be classified", molecular_subtype) ~ "GNG/GNT, Other alteration",
    !is.na(molecular_subtype) ~ molecular_subtype
  ))

subtypes <- hist %>%
  count(mol_sub_group) %>%
  dplyr::filter(n >= 5, !is.na(mol_sub_group)) %>%
  pull(mol_sub_group)

hist <- hist %>%
  dplyr::mutate(subtype = case_when(
    mol_sub_group %in% subtypes & !grepl("classified", mol_sub_group) ~ mol_sub_group,
    TRUE ~ "Other"
  ))

subtype_enr <- plp_enrichment(hist[hist$subtype != "Other",], "subtype")
subtype_enr <- subtype_enr %>%
  dplyr::mutate(subtype = glue::glue("{subtype} (N={cpgPLP + no_cpgPLP})")) %>%
  write_tsv(file.path(results_dir, "plp-carrier-subtype-counts.tsv"))

subtype_enr_plot <- subtype_enr %>% 
  dplyr::mutate(subtype = fct_relevel(subtype,
                                          rev(unique(subtype)))) %>%
  ggplot(aes(x = log10(OR), y = subtype, label = fdr_label)) +
  geom_point(size = 3, color = "#00A087FF",
             show.legend = FALSE) + 
  geom_errorbar(aes(xmin = log10(CI_lower), xmax = log10(CI_upper)), width = 0.2, 
                show.legend = FALSE, color = "#00A087FF") +
  geom_text(x = log10(subtype_enr$CI_upper)+0.25, hjust = 0, vjust = .5, size = 4, fontface = 2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "log10-OR (95% CI)", y = NULL) + 
  xlim(-2.5, 4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_Publication() +
  theme(plot.margin = unit(c(2,0.5,1,0.25), "lines"))

pdf(file.path(plot_dir, "subtype_plp_carrier_enr.pdf"),
    width = 5, height = 8)
subtype_enr_plot
dev.off()
```

For tumors categorized into "Other tumor" or "Other CNS embryonal tumor" plot_groups, calculate and plot P-LP enrichment among specific cancer group cohorts

```{r}
other_groups <- hist %>%
  dplyr::filter(plot_group %in% c("Other tumor", "Other CNS embryonal tumor")) %>%
  count(cancer_group) %>%
  dplyr::filter(n >= 5, !is.na(cancer_group)) %>%
  pull(cancer_group)

other_hist <- hist %>%
  dplyr::filter(plot_group %in% c("Other tumor", "Other CNS embryonal tumor"))

other_group_enr <- plp_enrichment(other_hist[other_hist$cancer_group != "Other tumor",], "cancer_group")

other_group_enr <- other_group_enr %>%
  dplyr::mutate(cancer_group = glue::glue("{cancer_group} (N={cpgPLP+no_cpgPLP})")) %>%
  write_tsv(file.path(results_dir, "plp-carrier-subtype-counts.tsv"))

other_group_enr_plot <- other_group_enr %>% 
  dplyr::mutate(cancer_group = fct_relevel(cancer_group,
                                          rev(unique(cancer_group)))) %>%
  ggplot(aes(x = log10(OR), y = cancer_group, label = fdr_label)) +
  geom_point(size = 3, color = "#00A087FF",
             show.legend = FALSE) + 
  geom_errorbar(aes(xmin = log10(CI_lower), xmax = log10(CI_upper)), width = 0.2, 
                show.legend = FALSE, color = "#00A087FF") +
  geom_text(x = log10(other_group_enr$CI_upper)+0.25, hjust = 0, vjust = .5, size = 4, fontface = 2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "log10-OR (95% CI)", y = NULL) + 
  xlim(-2.5, 4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  theme_Publication() +
  theme(plot.margin = unit(c(2,0.5,1,0.25), "lines"))

pdf(file.path(plot_dir, "other_group_plp_carrier_enr.pdf"),
    width = 6, height = 5)
other_group_enr_plot
dev.off()
```

```{r echo=TRUE}
sessionInfo()
```