---
title: "Plot HGG H3 Wt promoter methylation heatmap"
output: 
  html_notebook:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
---

This script generates a heatmap of HGG H3 wt samples by CpG promoter methylation 

# Load libraries
```{r load libraries, set up directories}
suppressPackageStartupMessages({
  library(ggplot2)
  library(tidyverse)
  library(ggpubr)
  library(ComplexHeatmap)
  library(circlize)
})

# Set up directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "methylation")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# If the directory does not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

# source publication theme
source(file.path(root_dir, "figures", "theme.R"))

# set seed to reproduce plots
set.seed(2023)
```


Set input file paths
```{r}
plp_file <- file.path(data_dir,
                      "pbta-merged-plp-variants-autogvp-abridged.tsv")

plp_sv_file <- file.path(data_dir,
                         "pbta_germline_svs.tsv")

histologies_file <- file.path(results_dir,
                              "germline-primary-plus-tumor-histologies-methylation.tsv")

promoter_methyl_file <- file.path(results_dir,
                                  "hgg-methyl-beta-value.tsv.gz")

methyl_annot_file <- file.path(root_dir, "data", "infinium.gencode.v39.probe.annotations.tsv.gz")

# cpg file
cpg_file <- file.path(root_dir, "analyses", "oncokb-annotation", "input", "cpg.txt")
```

Wrange data

```{r}
hist <- read_tsv(histologies_file)

methyl_annot <- read_tsv(methyl_annot_file)

promoter_probes <- methyl_annot %>%
  dplyr::filter(!Chromosome %in% c("chrX", "chrY"),
                Gene_Feature == "promoter") %>%
  distinct(Probe_ID) %>%
  pull(Probe_ID)

# filter matrix for promoter-annotated probes
methyl_mat <- read_tsv(promoter_methyl_file) %>%
  dplyr::filter(Probe_ID %in% promoter_probes) %>%
    column_to_rownames("Probe_ID")
```

Filter hist for HGG H3 wildtype samples with methylation data

```{r}
hist_methyl_hgg <- hist %>%
#  dplyr::filter(!is.na(Kids_First_Biospecimen_ID_methyl)) %>%
  dplyr::mutate(tp53_altered = case_when(
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("TP53", molecular_subtype) ~ "Yes",
    TRUE ~ "No"
  )) %>%
  dplyr::filter(Kids_First_Biospecimen_ID_methyl %in% colnames(methyl_mat)) %>%
  dplyr::arrange(factor(Kids_First_Biospecimen_ID_methyl),
                 levels = colnames(methyl_mat))
```

Load CPG P/LP variant information and append to hist

```{r}
cpgs <- read_lines(cpg_file)

plp <- read_tsv(plp_file) %>%
  dplyr::rename(Hugo_Symbol = gene_symbol_vep) %>%
  dplyr::mutate(variant_classification_vep = case_when(
    grepl("frameshift", variant_classification_vep) ~ "frameshift_variant",
    grepl("stop_gained", variant_classification_vep) ~ "stop_gained",
    grepl("missense_variant", variant_classification_vep) ~ "missense_variant",
    grepl("splice_donor|splice_acceptor", variant_classification_vep) ~ "splice_site",
    grepl("splice", variant_classification_vep) ~ "splice_region",
    TRUE ~ variant_classification_vep
  ))

plp_sv <- read_tsv(plp_sv_file) %>%
  dplyr::rename(Hugo_Symbol = Hugo_Symbol_cpg,
                variant_classification_vep = Type)

plp_summary <- plp %>%
  dplyr::filter(Hugo_Symbol %in% cpgs) %>%
  distinct(Kids_First_Biospecimen_ID_normal, Hugo_Symbol, variant_classification_vep) %>%
  bind_rows(plp_sv %>% dplyr::select(Kids_First_Biospecimen_ID_normal, 
                                     Hugo_Symbol, variant_classification_vep)) %>%
  group_by(Kids_First_Biospecimen_ID_normal) %>%
  summarise(Hugo_Symbol = str_c(Hugo_Symbol, collapse = ","),
            variant_classification_vep = str_c(variant_classification_vep, collapse = ",")) %>%
  dplyr::mutate(variant_classification_vep = case_when(
    grepl("stop_gained,stop_gained", variant_classification_vep) ~ "stop_gained",
    grepl("splice_site,splice_site", variant_classification_vep) ~ "splice_site",
    grepl(",", variant_classification_vep) ~ "multiple",
    TRUE ~ variant_classification_vep
  ))

hist_methyl_hgg <- hist_methyl_hgg %>%
  left_join(plp_summary) %>%
  dplyr::mutate(Hugo_Symbol = case_when(
    is.na(Hugo_Symbol) ~ "",
    TRUE ~ Hugo_Symbol
  ))
```

Load global methylation value df 

```{r}
global_methylation <- read_tsv(file.path(results_dir,
                                         "pbta-germline-mean-sample-methyl-beta.tsv")) %>%
  dplyr::filter(Kids_First_Biospecimen_ID_methyl %in% hist_methyl_hgg$Kids_First_Biospecimen_ID_methyl) %>%
  dplyr::arrange(factor(Kids_First_Biospecimen_ID_methyl), levels = hist_methyl_hgg$Kids_First_Biospecimen_ID_methyl)
```

Generate heatap annotation object

```{r}
gene_labels <- parse(text = paste0('italic("', hist_methyl_hgg$Hugo_Symbol, '")'))

ha_hgg = HeatmapAnnotation(
  `Tumor Descriptor` = hist_methyl_hgg$tumor_descriptor,
  `TP53 altered` = hist_methyl_hgg$tp53_altered,
  `Average methylation` = global_methylation$mean_beta_value,
  `P/LP variant classification` = hist_methyl_hgg$variant_classification_vep,
  Gene_symbol = anno_text(
    gene_labels,
    gp = gpar(fontsize = 9),
    just = "right"
  ),
 col = list("Tumor Descriptor" = c("Initial CNS Tumor" = "#F0E442",
                                   "Progressive" = "#56B4E9",
                                   "Deceased" = "#009E73",
                                   "Recurrence" = "#E69F00",
                                   "Second Malignancy" = "#0072B2"),
            "TP53 altered" = c("Yes" = "black", "No" = "whitesmoke"),
            "Average methylation" = colorRamp2(c(0.3, 0.5, 0.7), c("navy", "whitesmoke", "orangered")),
            "P/LP variant classification" = c("DEL" = "#313695" , "missense_variant" = "#3CB371","frameshift_variant" ="#FFBBFF","stop_gained" ="#6828E8", "splice_site" = "#F0E442", "multiple" = "#FF3E96", "splice_region" = "#7B68EE", "synonymous_variant" = "#FFE4E1")),
                  annotation_name_gp= gpar(fontsize = 10))

col_fun = colorRamp2(c(0, 0.5, 1), c("navy", "whitesmoke", "orangered"))
```

Calculate beta value variances and extract 20k most variable probes for plotting

```{r}
methyl_var <- apply(methyl_mat, 1, var)
names(methyl_var) <- rownames(methyl_mat)

var_probes <- names(sort(methyl_var, decreasing = T))[1:20000]
```

Plot heatmap

```{r}
ht_hgg <- Heatmap(methyl_mat[var_probes,hist_methyl_hgg$Kids_First_Biospecimen_ID_methyl],
                  col = col_fun,
              name = "Promoter CpG beta value",
              show_row_names = FALSE,
              show_column_names = FALSE,
              top_annotation = ha_hgg,
              use_raster = FALSE,
              heatmap_legend_param = list(legend_gp = gpar(fontsize = 10),
                                          labels_gp = gpar(fontsize = 10)))

pdf(file.path(plot_dir, "pbta-germline-hgg-wt-plp-somatic-prom-methylation.pdf"),
               width = 7.5, height = 10)

draw(ht_hgg, merge_legend = TRUE, heatmap_legend_side = "right",
     annotation_legend_side = "right")

dev.off()
```

Print session info

```{r}
sessionInfo()
```
