---
title: "Format PBTA-germline data for survival analyses"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2023
params:
  plot_ci: TRUE
---

**Purpose:** 

Format molecular subtype and event-free survival in PBTA-germline cohort for survival analyses

## Setup

#### Packages and functions

Load packages and set directory paths
```{r load packages, warning=FALSE}
library(tidyverse)
library(survival)
library(ggpubr)
library(ggplot2)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")

# If the input and results directories do not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}
```

Set file paths
```{r set paths}
histologies_file <- file.path(root_dir, "analyses", 
                              "collapse-tumor-histologies", "results", 
                              "germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

plp_file <- file.path(root_dir, "analyses", 
                      "variant-distribution", "input", 
                      "pbta_germline_plp_calls_autogvp_abridged_gnomad_popmax.tsv")

plp_sv_file <- file.path(root_dir, "analyses", 
                         "germline-sv", "input", 
                         "pbta_germline_svs.tsv")

cpg_file <- file.path(root_dir, "analyses", "oncokb-annotation", "input", "cpg.txt")

output_histology <- file.path(results_dir, "germline-primary-plus-tumor-histologies-plot-groups-clin-meta-subtype.tsv")
```

Import histologies, plp, and cpg files
```{r read files, echo=FALSE}
hist <- read_tsv(histologies_file)

plp <- read_tsv(plp_file, show_col_types = FALSE) 

plp_sv <- read_tsv(plp_sv_file, show_col_types = FALSE)

cpgs <- read_lines(cpg_file)

```

Consolidate subtypes of HGG, LGG, and GNG into broader subgroups
```{r}
hist <- hist %>%
    mutate(mol_sub_group = case_when(
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("wildtype", molecular_subtype) ~ "HGG, H3 WT",
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("K28", molecular_subtype) ~ "DMG, H3 K28",
    grepl("G35", molecular_subtype) ~ "DHG, H3 G35",
    grepl("IHG", molecular_subtype) ~ "IHG",
    plot_group %in% c("DIPG or DMG", "Other high-grade glioma") & grepl("IDH", molecular_subtype) ~ "HGG, IDH",
    plot_group %in% c("Low-grade glioma") & grepl("SEGA", molecular_subtype) ~ "SEGA",
    plot_group %in% c("Low-grade glioma") & grepl("BRAF V600E", molecular_subtype) ~ "LGG, BRAF V600E",
    plot_group %in% c("Low-grade glioma") & grepl("-BRAF", molecular_subtype) ~ "LGG, BRAF fusion",
    plot_group %in% c("Low-grade glioma") & grepl("wildtype", molecular_subtype) ~ "LGG, WT",
    plot_group %in% c("Low-grade glioma") & grepl("To be classified", molecular_subtype) ~ "LGG, To be classified",
    plot_group %in% c("Low-grade glioma") ~ "LGG, Other alteration",
    plot_group %in% c("Mixed neuronal-glial tumor") & grepl("wildtype", molecular_subtype) ~ "GNG/GNT, WT",
    plot_group %in% c("Mixed neuronal-glial tumor") & grepl("BRAF V600E", molecular_subtype) ~ "GNG/GNT, BRAF V600E",
    plot_group %in% c("Mixed neuronal-glial tumor") & grepl("-BRAF", molecular_subtype) ~ "GNG/GNT, BRAF fusion",
    plot_group %in% c("Mixed neuronal-glial tumor") & !grepl("To be classified", molecular_subtype) ~ "GNG/GNT, Other alteration",
    !is.na(molecular_subtype) ~ molecular_subtype
  ))
```

## Format survival data

Define column `cpgPLP_status` indicating presence or absence of a CPG PLP variant

```{r}
cpg_bs <- plp %>%
  filter(Kids_First_Biospecimen_ID_normal %in% hist$Kids_First_Biospecimen_ID_normal) %>%
  filter(gene_symbol_vep %in% cpgs) %>%
  pull(Kids_First_Biospecimen_ID_normal) %>%
  unique()

sv_cpg_bs <- plp_sv %>%
  pull(Kids_First_Biospecimen_ID_normal) %>%
  unique()

hist <- hist %>%
  mutate(cpgPLP_status = case_when(
    Kids_First_Biospecimen_ID_normal %in% c(cpg_bs, sv_cpg_bs) ~ "cpgPLP",
    TRUE ~ "no_cpgPLP"
  ))
```

We will define `EFS_status` as "NO EVENT" if `EFS_event_type` has values "Not Applicable" or "Not Reported", and as "EVENT" in all other cases. 

```{r}

hist <- hist %>%
  mutate(EFS_days = as.numeric(EFS_days),
         OS_days = as.numeric(OS_days)) %>%
  mutate(OS_years = OS_days/365.25,
         EFS_years = EFS_days/365.25) %>%
  dplyr::mutate(EFS_status = case_when(
    EFS_event_type %in% c("Not Applicable", "Not Reported") ~ "NO EVENT",
    !is.na(EFS_event_type) ~ "EVENT",
    TRUE ~ NA_character_
  ))

# Save output
write_tsv(hist,
          output_histology)
```
