---
title: "PBTA-germline survival analyses"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2023
params:
  plot_ci: TRUE
---

Run Kaplan-Meier survival analyses and Cox proportional hazards models on cancer group in the PBTA-germline cohort

## Usage 

Uses a wrapper function (`survival_analysis`) from utils folder. 

## Setup

#### Packages and functions

Load packages, set directory paths and call setup script

```{r load packages, warning=FALSE}
library(tidyverse)
library(survival)
library(ggpubr)
library(ggplot2)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")
input_dir <- file.path(analysis_dir, "results")
results_dir <- file.path(analysis_dir, "results")

# If the input and results directories do not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

source(file.path(analysis_dir, "util", "survival_models.R"))
```

Set metadata path

```{r set paths}
histologies_file <- file.path(input_dir, "germline-primary-plus-tumor-histologies-plot-groups-clin-meta-subtype.tsv")
```

Read in metadata

```{r}
hist <- read_tsv(histologies_file) %>%
  dplyr::mutate(age_at_diagnosis_days = as.numeric(age_at_diagnosis_days)) %>%
  dplyr::mutate(age_at_diagnosis_years = age_at_diagnosis_days/365.25) %>%
  dplyr::mutate(cpgPLP_status = fct_relevel(cpgPLP_status, 
                                            c("no_cpgPLP",
                                              "cpgPLP"))) %>%
  dplyr::mutate(extent_of_tumor_resection = case_when(
    extent_of_tumor_resection == "Biopsy only" & plot_group == "Medulloblastoma" ~ "Partial resection",
    TRUE ~ extent_of_tumor_resection
  )) %>%
  dplyr::mutate(extent_of_tumor_resection = fct_relevel(extent_of_tumor_resection,
                                                        c("Gross/Near total resection",
                                                          "Partial resection", "Biopsy only",
                                                          "Unavailable")))
```

Define plot groups for which survival analyses should be run

```{r}
groups <- c("Atypical Teratoid Rhabdoid Tumor",
            "Choroid plexus tumor",
            "Craniopharyngioma",
            "DIPG or DMG",
            "Other high-grade glioma",
            "Ependymoma",
            "Low-grade glioma",
            "Medulloblastoma",
            "Meningioma",
            "Mesenchymal tumor",
            "Mixed neuronal-glial tumor",
            "Neurofibroma plexiform"
            )

# create names vector for output files
plot_names <- c("ATRT", "CPT", "CRANIO", "DMG", "HGG", "EPN", "LGG", "MB", "MNG", "MES", "GNG-GNT", "NFP")
names(plot_names) <- groups
```

Loop through groups and create Kaplan Meier survival and Cox Proportional Hazard models for overall and event-free survival

```{r}
for (group in groups){
  
  results_dir <- file.path(analysis_dir, "results", plot_names[group])
    
  if (!dir.exists(results_dir)) {
      dir.create(results_dir)
      
    }
  
  # Define output model files
  km_os_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{plot_names[group]}_OS_cpgPLPstatus.RDS"))
  
  coxph_os_model_file <- ifelse(group %in% c("Low-grade glioma"),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_resection_cpgPLPstatus.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_OS_additive_terms_subtype_cpgPLPstatus.RDS"))
  )
  
  km_efs_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{plot_names[group]}_EFS_cpgPLPstatus.RDS"))
  
  coxph_efs_model_file <- ifelse(group %in% c("Low-grade glioma"),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_resection_cpgPLPstatus.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{plot_names[group]}_EFS_additive_terms_subtype_cpgPLPstatus.RDS"))
  )
  
  # Subset histology
  group_hist <- hist %>%
    filter(grepl(group, plot_group))
  
  if (group == "Other high-grade glioma"){
    group_hist <- group_hist %>%
       dplyr::mutate(molecular_subtype = fct_relevel(molecular_subtype,
                                      c("HGG, H3 wildtype", "HGG, H3 wildtype, TP53",
                                        "DHG, H3 G35", "DHG, H3 G35, TP53",
                                        "HGG, IDH", "HGG, IDH, TP53",
                                        "IHG, ALK-altered", "IHG, NTRK-altered",
                                        "IHG, NTRK-altered, TP53", 
                                        "IHG, ROS1-altered",
                                        "HGG, PXA"))) %>%
     arrange(molecular_subtype)
  }
  
    if (group == "Ependymoma") {
      group_hist <- group_hist %>%
        dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                      c("EPN, ST ZFTA", "EPN, ST YAP1", "EPN, PF A", 
                        "EPN, PF B", "EPN, MPE", "EPN, SP", "EPN, SP-MYCN",
                        "EPN, To be classified")))
        
    } 
    
    if (group == "Low-grade glioma") {
      group_hist <- group_hist %>%
        dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                                                  c("LGG, WT", 
                                                    "LGG, BRAF V600E",
                                                    "LGG, BRAF fusion",
                                                    "LGG, Other alteration",
                                                    "SEGA",
                                                    "LGG, To be classified")))
    }
    
    if (group == "Mixed neuronal-glial tumor") {
      group_hist <- group_hist %>%
        dplyr::mutate(mol_sub_group = fct_relevel(mol_sub_group,
                                                  c("GNG/GNT, WT", 
                                                    "GNG/GNT, BRAF V600E",
                                                    "GNG/GNT, BRAF fusion",
                                                    "GNG/GNT, Other alteration",
                                                    "GNT, To be classified")))
    }
  
   
  # generate kaplan-meier survival models and save
  kap_os <- survival_analysis(
  metadata  = group_hist,
  ind_var = "cpgPLP_status",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal",
  days_col = "OS_days",
  status_col = "OS_status"
  )

  readr::write_rds(kap_os, km_os_model_file)
  
  kap_efs <- survival_analysis(
  metadata  = group_hist,
  ind_var = "cpgPLP_status",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal",
  days_col = "EFS_days",
  status_col = "EFS_status"
  )

  readr::write_rds(kap_efs, km_efs_model_file)


  # generate coxph models (including appropriate covariates) and save
  add_model_os <- fit_save_model(group_hist[!grepl("To be classified", group_hist$mol_sub_group) & !grepl("Unavailable", group_hist$extent_of_tumor_resection),],
                                  terms = ifelse(
                                     group %in% c("Low-grade glioma"), 
                                     "mol_sub_group+extent_of_tumor_resection+cpgPLP_status",
                                     ifelse(group %in% c("Other high-grade glioma"),
                                            "extent_of_tumor_resection+molecular_subtype+cpgPLP_status",
                                            ifelse(group == "DIPG or DMG",
                                                   "molecular_subtype+cpgPLP_status",
                                                   ifelse(group %in% c("Atypical Teratoid Rhabdoid Tumor", "Medulloblastoma"),
                                                          "extent_of_tumor_resection+mol_sub_group+cpgPLP_status",
                                                          ifelse(group %in% c("Mixed neuronal-glial tumor", "Ependymom"),
                                                                 "mol_sub_group+cpgPLP_status",
                                                                 ifelse(group %in% c("Choroid plexus tumor", "Mesenchymal tumor",
                                                                                     "Meningioma"),
                                                                        "extent_of_tumor_resection+cpgPLP_status",
                                                                        "cpgPLP_status"
                                                                        )))))),
             coxph_os_model_file,
             "multivariate",
             years_col = "OS_years",
             status_col = "OS_status"
            )
  
  add_model_efs <- fit_save_model(group_hist[!grepl("To be classified", group_hist$mol_sub_group) & !grepl("Unavailable", group_hist$extent_of_tumor_resection),],
                                  terms = ifelse(
                                     group %in% c("Low-grade glioma"), 
                                     "mol_sub_group+extent_of_tumor_resection+cpgPLP_status",
                                     ifelse(group %in% c("Other high-grade glioma"),
                                            "extent_of_tumor_resection+molecular_subtype+cpgPLP_status",
                                            ifelse(group == "DIPG or DMG",
                                                   "molecular_subtype+cpgPLP_status",
                                                   ifelse(group %in% c("Atypical Teratoid Rhabdoid Tumor", "Medulloblastoma"),
                                                          "extent_of_tumor_resection+mol_sub_group+cpgPLP_status",
                                                          ifelse(group %in% c("Mixed neuronal-glial tumor", "Ependymom"),
                                                                 "mol_sub_group+cpgPLP_status",
                                                                 ifelse(group %in% c("Choroid plexus tumor", "Mesenchymal tumor",
                                                                                     "Meningioma"),                                                                        
                                                                        "extent_of_tumor_resection+cpgPLP_status",
                                                                        "cpgPLP_status"
                                                                        )))))),
             coxph_efs_model_file,
             "multivariate",
             years_col = "EFS_years",
             status_col = "EFS_status"
            )
  
}
```

Loop through subtypes with sufficient sample size and build KM and coxph survival models

```{r}
# define subtype df with histology, subtype, and file names
subtypes <- hist %>%
  count(mol_sub_group, plot_group) %>%
  filter(n >=15 & !grepl("classified", mol_sub_group) & !is.na(mol_sub_group)) %>%
  dplyr::mutate(hist = unlist(lapply(strsplit(mol_sub_group, ", "), function(x) x[1])),
                subtype = case_when(
                  grepl(",", mol_sub_group) ~ unlist(lapply(strsplit(mol_sub_group, ", "), function(x) x[2])),
                  TRUE ~ NA_character_)) %>%
  dplyr::mutate(hist = case_when(
    grepl("GNG/GNT", hist) ~ "GNG-GNT",
    TRUE ~ hist
  )) %>%
  dplyr::mutate(subtype = str_replace_all(subtype, " ", "-")) %>%
  write_tsv(file.path(analysis_dir, "results", "subtypes-for-survival.tsv"))


# Loop through subtypes and run survival analyses
for (j in 1:nrow(subtypes)) {
  
    results_dir <- file.path(analysis_dir, "results", subtypes$hist[j])
    
    if (!dir.exists(results_dir)) {
      dir.create(results_dir)
      
    }
  
  # define output files
  km_os_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{subtypes$hist[j]}_{subtypes$subtype[j]}_OS_cpgPLPstatus.RDS"))
  
  coxph_os_model_file <- ifelse(subtypes$plot_group[j] %in% c("Low-grade glioma"),
                             file.path(results_dir,
                                      glue::glue("cox_{subtypes$hist[j]}_{subtypes$subtype[j]}_OS_additive_terms_subtype_resection_cpgPLPstatus.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{subtypes$hist[j]}_{subtypes$subtype[j]}_OS_additive_terms_subtype_cpgPLPstatus.RDS"))
  )
  
  km_efs_model_file <- file.path(results_dir, 
                                    glue::glue("logrank_{subtypes$hist[j]}_{subtypes$subtype[j]}_EFS_cpgPLPstatus.RDS"))
  
  coxph_efs_model_file <- ifelse(subtypes$plot_group[j] %in% c("Low-grade glioma"),
                             file.path(results_dir,
                                       glue::glue("cox_{subtypes$hist[j]}_{subtypes$subtype[j]}_EFS_additive_terms_subtype_resection_cpgPLPstatus.RDS")),
                             file.path(results_dir,
                                       glue::glue("cox_{subtypes$hist[j]}_{subtypes$subtype[j]}_EFS_additive_terms_subtype_cpgPLPstatus.RDS"))
  )
  
  
  # Subset metadata for histology subtype
  subtype_hist <- hist %>%
    filter(grepl(subtypes$plot_group[j], plot_group) & grepl(subtypes$mol_sub_group[j], mol_sub_group))
  
  
  # Run kaplan meier survival models and save to output
  kap_os <- survival_analysis(
              metadata  = subtype_hist,
              ind_var = "cpgPLP_status",
              test = "kap.meier",
              metadata_sample_col = "Kids_First_Biospecimen_ID_normal",
              days_col = "OS_days",
              status_col = "OS_status"
              )

  readr::write_rds(kap_os, 
                   str_replace_all(km_os_model_file, " ", "-"))
  
  kap_efs <- survival_analysis(
                metadata  = subtype_hist,
                ind_var = "cpgPLP_status",
                test = "kap.meier",
                metadata_sample_col = "Kids_First_Biospecimen_ID_normal",
                days_col = "EFS_days",
                status_col = "EFS_status"
                )
  
    readr::write_rds(kap_efs, 
                   str_replace_all(km_efs_model_file, " ", "-"))


    # Run coxph models and save to output
    add_model_os <- fit_save_model(subtype_hist[!grepl("Unavailable", subtype_hist$extent_of_tumor_resection),],
                                 terms = ifelse(subtypes$plot_group[j] %in% c("Low-grade glioma",
                                                                               "Medulloblastoma"),
                                                 "extent_of_tumor_resection+cpgPLP_status",
                                                 ifelse(subtypes$plot_group[j] %in% c("Other high-grade glioma"),
                                                        "extent_of_tumor_resection+molecular_subtype+cpgPLP_status",
                                                        ifelse(subtypes$plot_group[j] %in% c("DIPG or DMG"),
                                                               "molecular_subtype+cpgPLP_status",
                                                               "cpgPLP_status"
                                                               ))),
                                   coxph_os_model_file,
                                   "multivariate",
                                   years_col = "OS_years",
                                   status_col = "OS_status")
    
    add_model_efs <- fit_save_model(subtype_hist[!grepl("Unavailable", subtype_hist$extent_of_tumor_resection),],
                                  terms = ifelse(subtypes$plot_group[j] %in% c("Low-grade glioma",
                                                                               "Medulloblastoma"),
                                                 "extent_of_tumor_resection+cpgPLP_status",
                                                 ifelse(subtypes$plot_group[j] %in% c("Other high-grade glioma"),
                                                        "extent_of_tumor_resection+molecular_subtype+cpgPLP_status",
                                                        ifelse(subtypes$plot_group[j] %in% c("DIPG or DMG"),
                                                               "molecular_subtype+cpgPLP_status",
                                                               "cpgPLP_status"
                                                               ))),
                                     coxph_efs_model_file,
                                     "multivariate",
                                     years_col = "EFS_years",
                                     status_col = "EFS_status"
                                    )

}
```

```{r}
sessionInfo()
```
