---
title: "PBTA-germline pathway-level survival analyses"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2023
params:
  plot_ci: TRUE
---

## Purpose

Determine the effect of germline DNA repair gene PLP variants on overall and event-free survival 

## Setup

#### Packages and functions

Load packages, set directory paths and call setup script
```{r}
library(tidyverse)
library(survival)
library(patchwork)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")

source(file.path(analysis_dir, "util", "survival_models.R"))

# Declare output directory
results_dir <- file.path(analysis_dir, "results")
plots_dir <- file.path(analysis_dir, "plots")

# Input directory
input_dir <- file.path(root_dir, "analyses", "survival", "results")
```

Set file paths
```{r set paths}
cbtn_histologies_file <- file.path(results_dir, "germline-primary-plus-tumor-histologies-plot-groups-clin-meta-subtype.tsv")

all_plp_file <- file.path(data_dir, "pbta_germline_plp_calls.tsv")

cpg_file <- file.path(root_dir, "analyses", "oncokb-annotation", "input", "cpg.txt")

km_mmr_output_pdf <- file.path(plots_dir, "MMR_km_survival_hgg.pdf")

km_brca_output_pdf <- file.path(plots_dir, "BRCA_km_survival_hgg.pdf")
```

Import data
```{r data wrangling}
hist <- read_tsv(cbtn_histologies_file)

all_plp <- read_tsv(all_plp_file)

cpgs <- read_lines(cpg_file)
```

Load mismatch repair (MMR) gene list and identify samples with germline P-LP variants in these genes
```{r}
mmr_genes <- read_lines(file.path(root_dir, "analyses", "dna-repair-variant-summary", "input", "Knijnenburg_paper", "mismatch_repair.txt"))

mmr_bs <- all_plp %>%
      filter(Hugo_Symbol %in% mmr_genes & Kids_First_Biospecimen_ID %in% hist$Kids_First_Biospecimen_ID_normal) %>%
      pull(Kids_First_Biospecimen_ID) %>%
      unique()
```

Subset histologies for HGG tumors and create column indicating presence/absence of MMR germline variant 
```{r}
hgg_hist <- hist %>%
  filter(plot_group %in% c("DIPG or DMG", "Other high-grade glioma")) %>%
  mutate(age_at_diagnosis_days = as.numeric(age_at_diagnosis_days)) %>%
  mutate(age_at_diagnosis_years = age_at_diagnosis_days/365.25) %>%
  mutate(MMR_variant = case_when(
    Kids_First_Biospecimen_ID_normal %in% mmr_bs ~ "MMR PLP",
    TRUE ~ "MMR WT"
  ))  %>%
  mutate(mol_sub_group = factor(mol_sub_group,
                                c("HGG, H3 WT", "DHG, H3 G35",
                                  "DMG, H3 K28", "HGG, IDH", "IHG"))) %>%
  mutate(MMR_variant = factor(MMR_variant,
                                   c("MMR WT",
                                     "MMR PLP")))
```

Print N per MMR variant status group
```{r}
table(hgg_hist$MMR_variant)

```
  
Run Kaplain meier tests for survival by MMR germline variant status and plot curves
```{r}
kap_os <- survival_analysis(
  metadata  = hgg_hist,
  ind_var = "MMR_variant",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal", 
  days_col = "OS_days",
  status_col = "OS_status"
  )

kap_efs <- survival_analysis(
  metadata  = hgg_hist,
  ind_var = "MMR_variant",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal", 
  days_col = "EFS_days",
  status_col = "EFS_status"
  )

km_plot <- plotKM(model = list(kap_os, kap_efs),
                    variable = "MMR_variant",
                    combined = T, 
                    title = "HGG")

ggsave(km_mmr_output_pdf, km_plot,
       width = 10, height = 6, units = "in",
       device = "pdf")
```



Define vector of BRCA/BRCA-interacting genes
```{r}
brca_bs <- all_plp %>%
      dplyr::filter(grepl("BRCA|BARD1|BRIP1|ATM|CHEK2|RAD51|PALB2", Hugo_Symbol) & !grepl("RAD51C", Hugo_Symbol)) %>%
      pull(Kids_First_Biospecimen_ID) %>%
      unique()
```

Create column indicating presence/absence of BRCA variant
```{r}
hgg_hist <- hgg_hist %>%
  mutate(brca_variant = case_when(
    Kids_First_Biospecimen_ID_normal %in% brca_bs ~ "BRCA PLP",
    TRUE ~ "BRCA WT"
  ))  %>%
  mutate(brca_variant = factor(brca_variant,
                                   c("BRCA WT",
                                     "BRCA PLP")))
```

Print N per BRCA variant status group
```{r}
table(hgg_hist$brca_variant)

```
  
Run KM survival models by BRCA variant status, and plot results
```{r}
kap_os <- survival_analysis(
  metadata  = hgg_hist,
  ind_var = "brca_variant",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal", 
  days_col = "OS_days",
  status_col = "OS_status"
  )

kap_efs <- survival_analysis(
  metadata  = hgg_hist,
  ind_var = "brca_variant",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal", 
  days_col = "EFS_days",
  status_col = "EFS_status"
  )

km_plot <- plotKM(model = list(kap_os, kap_efs),
                    variable = "brca_variant",
                    combined = T, 
                    title = "HGG")

ggsave(km_brca_output_pdf, km_plot,
       width = 10, height = 6, units = "in",
       device = "pdf")
```

Run cox proportional hazards models for HGG OS and EFS, including covariates `mol_sub_group`, `age_at_diagnosis_years`, `MMR_variant`, and `brca_variant`

Interaction OS model: 
```{r coxph int OS repair plp}  

int_os_model <- fit_save_model(hgg_hist,
                             "age_at_diagnosis_years+mol_sub_group*MMR_variant+mol_sub_group*brca_variant",
                                 file.path(results_dir, "coxph_int_hgg_OS_subtype_age_repairPLP.RDS"),
                                 "multivariate",
                                 years_col = "OS_years",
                                 status_col = "OS_status")

int_os_forest_plot <- plotForest(readRDS(file.path(results_dir, "coxph_int_hgg_OS_subtype_age_repairPLP.RDS")))

ggsave(file.path(plots_dir, "forest_hgg_int_OS_subtype_age_repairPLP.pdf"), 
       int_os_forest_plot,
       width = 12, height = 3, units = "in",
       device = "pdf")

readRDS(file.path(results_dir, "coxph_int_hgg_OS_subtype_age_repairPLP.RDS"))
```
Additive OS model: 
```{r coxph add OS repair plp}
add_os_model <- fit_save_model(hgg_hist,
                             "mol_sub_group+age_at_diagnosis_years+MMR_variant+brca_variant",
                                 file.path(results_dir, "coxph_add_hgg_OS_subtype_age_repairPLP.RDS"),
                                 "multivariate",
                                 years_col = "OS_years",
                                 status_col = "OS_status")

add_os_forest_plot <- plotForest(readRDS(file.path(results_dir, "coxph_add_hgg_OS_subtype_age_repairPLP.RDS")))

ggsave(file.path(plots_dir, "forest_hgg_add_OS_subtype_age_repairPLP.pdf"), 
       add_os_forest_plot,
       width = 8, height = 3, units = "in",
       device = "pdf")

readRDS(file.path(results_dir, "coxph_add_hgg_OS_subtype_age_repairPLP.RDS"))
```

Interaction EFS model: 
```{r coxph int EFS repair plp}
int_efs_model <- fit_save_model(hgg_hist,
                             "age_at_diagnosis_years+mol_sub_group*MMR_variant+mol_sub_group*brca_variant",
                             file.path(results_dir, "coxph_int_hgg_EFS_subtype_age_repairPLP.RDS"),
                             "multivariate",
                             years_col = "EFS_years",
                             status_col = "EFS_status")

int_efs_forest_plot <- plotForest(readRDS(file.path(results_dir, "coxph_int_hgg_EFS_subtype_age_repairPLP.RDS")))

ggsave(file.path(plots_dir, "forest_hgg_int_EFS_subtype_age_repairPLP.pdf"), 
       int_efs_forest_plot,
       width = 12, height = 3, units = "in",
       device = "pdf")

readRDS(file.path(results_dir, "coxph_int_hgg_EFS_subtype_age_repairPLP.RDS"))

```

Additive EFS model: 
```{r coxph add EFS repair plp}
add_efs_model <- fit_save_model(hgg_hist,
                             "mol_sub_group+age_at_diagnosis_years+MMR_variant+brca_variant",
                             file.path(results_dir, "coxph_add_hgg_EFS_subtype_age_repairPLP.RDS"),
                             "multivariate",
                             years_col = "EFS_years",
                             status_col = "EFS_status")

efs_forest_plot <- plotForest(readRDS(file.path(results_dir, "coxph_add_hgg_EFS_subtype_age_repairPLP.RDS")))

ggsave(file.path(plots_dir, "forest_hgg_add_EFS_subtype_age_repairPLP.pdf"), 
       efs_forest_plot,
       width = 8, height = 3, units = "in",
       device = "pdf")

readRDS(file.path(results_dir, "coxph_add_hgg_EFS_subtype_age_repairPLP.RDS"))

```

```{r}
sessionInfo()

```