---
title: "Assess P/LP carrier status among LGG BRAF fusion breakpoint groups"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2023
params:
  plot_ci: TRUE
---

**Purpose:** 

Assess distribution of KIAA1549:BRAF fusion breakpoints among germline cancer predisposition gene P/LP variant carriers in PBTA germline cohort

## Setup

#### Packages and functions

Load packages and set directory paths
```{r load packages, warning=FALSE}
library(tidyverse)
library(survival)
library(ggpubr)
library(ggplot2)
library(circlize)
library(ComplexHeatmap)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# If the input and results directories do not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

source(file.path(analysis_dir, "util",
                 "survival_models.R"))
source(file.path(root_dir, "analyses", 
                  "demo-clin-stats", "util",
                 "heatmap_function.R"))
```

Set file paths

```{r set paths}
histologies_file <- file.path(results_dir,
                              "germline-primary-plus-tumor-histologies-plot-groups-clin-meta-subtype.tsv")

plp_file <- file.path(root_dir, "analyses", "two-hits", "results", 
                      "germline-plp-variants-cpg-somatic-driver-in-tumor.tsv")

plp_sv_file <- file.path(data_dir,
                         "pbta_germline_svs.tsv")

cpg_file <- file.path(root_dir, "analyses", "oncokb-annotation", "input", "cpg.txt")

fusions_file <- file.path(input_dir, "lgg-braf-fusion-breakpoint-annotation.tsv")

cns_region_file <- file.path(input_dir, "pbta-lgg-braf-fusion-mixed-location.tsv")
```

Wrangle data

```{r}
hist <- read_tsv(histologies_file, guess_max = 10000) %>%
  dplyr::mutate(age_at_diagnosis_years = age_at_diagnosis_days/365.25)

cpgs <- read_lines(cpg_file)

plp <- read_tsv(plp_file) %>%
  dplyr::filter(Hugo_Symbol %in% cpgs)

plp_sv <- read_tsv(plp_sv_file)

cns_regions <- read_tsv(cns_region_file)
```

Obtain all BS IDs of P/LP carriers with BRAF fusion-positive LGG

```{r}
plp_ids <- unique(c(plp$Kids_First_Biospecimen_ID_normal,
                    plp_sv$Kids_First_Biospecimen_ID_normal))

braf_plp <- hist %>%
  dplyr::filter(plot_group == "Low-grade glioma" & grepl("BRAF fusion", mol_sub_group),
                Kids_First_Biospecimen_ID_normal %in% plp_ids) %>%
  pull(Kids_First_Participant_ID)
```

Load braf fusions file, and add column indicating if patient is CPG P/LP carrier

```{r}
fusions <- read_tsv(fusions_file) %>%
  dplyr::rename(path_diagnosis = diagnosis) %>%
  dplyr::arrange(path_diagnosis) %>%
  distinct(Kids_First_Participant_ID, .keep_all = TRUE) %>%
  dplyr::filter(Kids_First_Participant_ID %in% hist$Kids_First_Participant_ID[hist$plot_group == "Low-grade glioma" & grepl("BRAF fusion", hist$mol_sub_group)]) %>%
  # define CPG P/LP status
  dplyr::mutate(cpg_plp = case_when(
    Kids_First_Participant_ID %in% braf_plp ~ "CPG P/LP",
    TRUE ~ "No CPG P/LP")) %>%
  # reorder breakpoint group levels
  dplyr::mutate(breakpoint_group = fct_relevel(breakpoint_group,
                                               c("16:09", "15:09", "16:11", "18:10", "rare"))) %>%
  dplyr::mutate(cpg_plp = fct_relevel(cpg_plp,
                                               c("No CPG P/LP", "CPG P/LP"))) %>%
  # Define CNS regions resolved after path reviews
  dplyr::mutate(CNS_region = case_when(
    Kids_First_Participant_ID %in% cns_regions$Kids_First_Participant_ID ~ cns_regions$CNS_region[match(Kids_First_Participant_ID, cns_regions$Kids_First_Participant_ID)],
    TRUE ~ CNS_region
  )) %>%
  # add survival and age at dx data
  left_join(hist %>% dplyr::select(Kids_First_Participant_ID,
                                   Kids_First_Biospecimen_ID_normal,
                                   OS_years, EFS_years,
                                   EFS_status, age_at_diagnosis_years)) %>%
  # reorder path diagnosis
  dplyr::mutate(path_diagnosis = fct_relevel(path_diagnosis,
                                        c("pilocytic astrocytoma",
                                          "pilomyxoid astrocytoma",
                                          "fibrillary astrocytoma"))) %>%
  # two samples with unreported tumor resection had path reports manually reviewed and confirmed resection levels. add here:
  dplyr::mutate(extent_of_tumor_resection = case_when(
    sample_id == "7316-436" ~ "Gross/Near total resection",
    sample_id == "7316-6762" ~ "Partial resection",
    TRUE ~ extent_of_tumor_resection
  )) %>%
  dplyr::mutate(extent_of_tumor_resection = fct_relevel(extent_of_tumor_resection,
                                                        c("Gross/Near total resection",
                                                          "Partial resection",
                                                          "Biopsy only")))

table(fusions$cpg_plp)
table(fusions$breakpoint_group, fusions$cpg_plp)
```
Survival model including tumor resection and age at dx additive effects, and diagnosis-P/LP status interaction effect 

```{r}
int_braf_fusions_plp_efs_model <- fit_save_model(fusions,
                                    "extent_of_tumor_resection+path_diagnosis*cpg_plp+age_at_diagnosis_years",
                                    file.path(results_dir, "coxph_EFS_lgg_braf_fusion_resection_age_int_diagnosis_plp.RDS"),
                                    "multivariate",
                                    years_col = "EFS_years",
                                    status_col = "EFS_status")

plotForest(readRDS(file.path(results_dir, "coxph_EFS_lgg_braf_fusion_resection_age_int_diagnosis_plp.RDS")))

ggsave(file.path(plot_dir, "forest_EFS_lgg_braf_fusion_resection_age_int_diagnosis_plp.pdf"),
       width = 7.5, height = 3, units = "in")
```

Pilocytic astrocytoma additive EFS model

```{r}
pa_fusions <- fusions %>%
  dplyr::filter(path_diagnosis == "pilocytic astrocytoma")

add_pa_plp_efs_model <- fit_save_model(pa_fusions,
                                    "extent_of_tumor_resection+cpg_plp+age_at_diagnosis_years",
                                    file.path(results_dir, "coxph_EFS_pa_resection_age_plp.RDS"),
                                    "multivariate",
                                    years_col = "EFS_years",
                                    status_col = "EFS_status")

plotForest(readRDS(file.path(results_dir, "coxph_EFS_pa_resection_age_plp.RDS")))

ggsave(file.path(plot_dir, "forest_EFS_pa_resection_age_plp.pdf"),
       width = 7, height = 3, units = "in")
```


Determine if breakpoint groups are enriched among P/LP carriers with PA

```{r}
fusion_enr <- plot_enr(pa_fusions, 
                       var1 = "breakpoint_group", 
                       var2 = "cpg_plp",
                       var1_names = c("15:09", "16:09", "16:11", "18:10", "rare"), 
                       var2_names = c("No CPG P/LP", "CPG P/LP"))

pdf(file.path(plot_dir, "pa_plp_carrier_fusion_enr_heatmap.pdf"),
    width = 3, height = 4)
draw(fusion_enr)
dev.off()
```

Assess PA extent of surgical resection by CPG P/LP status

```{r}
pa_plp_resection_enr <- plot_enr(pa_fusions, 
                       var2 = "cpg_plp", 
                       var1 = "extent_of_tumor_resection",
                       var2_names = c("No CPG P/LP", "CPG P/LP"), 
                       var1_names = unique(pa_fusions$extent_of_tumor_resection),
                       padjust = TRUE)

pdf(file.path(plot_dir, "pa_cpg_plp_resection_enr_heatmap.pdf"),
    width = 5.5, height = 3)
draw(pa_plp_resection_enr)
dev.off()
```

Assess PA CNS region by CPG P/LP status

```{r}
pa_plp_region_enr <- plot_enr(pa_fusions, 
                       var2 = "cpg_plp", 
                       var1 = "CNS_region",
                       var2_names = c("No CPG P/LP", "CPG P/LP"), 
                       var1_names = unique(pa_fusions$CNS_region),
                       padjust = TRUE)

pdf(file.path(plot_dir, "pa_cpg_plp_region_enr_heatmap.pdf"),
    width = 5.5, height = 3)
draw(pa_plp_region_enr)
dev.off()
```

Print P/LP variants in patients with PA

```{r}
plp %>%
  dplyr::select(Kids_First_Biospecimen_ID_normal, Hugo_Symbol) %>%
  bind_rows(plp_sv %>% 
              dplyr::rename(Hugo_Symbol = Hugo_Symbol_cpg) %>%
              dplyr::select(Kids_First_Biospecimen_ID_normal, Hugo_Symbol)) %>%
  left_join(hist %>% dplyr::select(Kids_First_Biospecimen_ID_normal, Kids_First_Participant_ID, EFS_days, EFS_status)) %>%
  dplyr::filter(Kids_First_Participant_ID %in% pa_fusions$Kids_First_Participant_ID)
```

```{r}
sessionInfo()
```