---
title: "Assess P/LP carrier status among MB cohort and associations with survival"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
author: Ryan Corbett
date: 2024
params:
  plot_ci: TRUE
---

**Purpose:** 

Assess survival in medulloblastoma cohort by CPG P/LP status, methylation subtypes, and metastasis status

## Setup

#### Packages and functions

Load packages and set directory paths
```{r load packages, warning=FALSE}
library(tidyverse)
library(survival)
library(ggpubr)
library(ggplot2)
library(circlize)
library(ComplexHeatmap)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "survival")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# If the input and results directories do not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

source(file.path(analysis_dir, "util",
                 "survival_models.R"))
source(file.path(root_dir, "analyses", 
                  "demo-clin-stats", "util",
                 "heatmap_function.R"))
```

Set file paths

```{r set paths}
mb_hist_file <- file.path(results_dir, 
                          "mb-histologies-plus-shh-methyl-subtypes-metastsis.tsv")
```

Wrangle data

```{r}
mb_hist <- read_tsv(mb_hist_file) %>%
  dplyr::mutate(metastasis = fct_relevel(metastasis, 
                                         c("No metastasis", "Metastasis"))) %>%
  dplyr::mutate(cpg_plp = fct_relevel(cpg_plp,
                                      c("No P/LP", "CPG P/LP"))) %>%
  dplyr::mutate(shh_type = fct_relevel(shh_type,
                                       "MB, child-adult SHH", "MB, infant SHH"))
```

Generate OS and EFS coxph models, adding covariates for metastasis status

```{r}
add_mb_efs_model <- fit_save_model(mb_hist[!grepl("classified", mb_hist$molecular_subtype),],
                                "extent_of_tumor_resection+molecular_subtype+metastasis+cpg_plp",
                                file.path(results_dir, "coxph_add_EFS_mb_resection_subtype_metastasis_plp.RDS"),
                                "multivariate",
                                years_col = "EFS_years",
                                status_col = "EFS_status")

plotForest(readRDS(file.path(results_dir, "coxph_add_EFS_mb_resection_subtype_metastasis_plp.RDS")))

ggsave(file.path(plot_dir, "forest_EFS_mb_resection_subtype_metastasis_plp.pdf"),
       width = 7, height = 3, units = "in")
```

```{r}
add_mb_os_model <- fit_save_model(mb_hist[!grepl("classified", mb_hist$molecular_subtype),],
                                "extent_of_tumor_resection+molecular_subtype+metastasis+cpg_plp",
                                file.path(results_dir, "coxph_add_OS_mb_resection_subtype_metastasis_plp.RDS"),
                                "multivariate",
                                years_col = "OS_years",
                                status_col = "OS_status")

plotForest(readRDS(file.path(results_dir, "coxph_add_OS_mb_resection_subtype_metastasis_plp.RDS")))

ggsave(file.path(plot_dir, "forest_OS_mb_resection_subtype_metastasis_plp.pdf"),
       width = 7, height = 3, units = "in")
```

Generate OS and EFS survival models by molecular subtype, including covariates for extent of tumor resection, metastasis, and cpg P/LP status

```{r}
subtypes <- unique(mb_hist$molecular_subtype[!grepl("classified", mb_hist$molecular_subtype)])

for (subtype in subtypes){
  
  abbrev <- str_replace(subtype, "MB, ", "")
  
  add_mb_efs_model <- fit_save_model(mb_hist[mb_hist$molecular_subtype == subtype,],
                                "extent_of_tumor_resection+metastasis+cpg_plp",
                                file.path(results_dir, glue::glue("coxph_add_EFS_mb_{abbrev}_resection_metastasis_plp.RDS")),
                                "multivariate",
                                years_col = "EFS_years",
                                status_col = "EFS_status")
  
  plotForest(readRDS(file.path(results_dir, glue::glue("coxph_add_EFS_mb_{abbrev}_resection_metastasis_plp.RDS"))))

  ggsave(file.path(plot_dir, glue::glue("forest_EFS_mb_{abbrev}_resection_subtype_metastasis.pdf")),
         width = 7, height = 3, units = "in")
  
  if (subtype != "MB, WNT"){
  
    add_mb_os_model <- fit_save_model(mb_hist[mb_hist$molecular_subtype == subtype,],
                                "extent_of_tumor_resection+metastasis+cpg_plp",
                                file.path(results_dir, glue::glue("coxph_add_OS_mb_{abbrev}_resection_metastasis_plp.RDS")),
                                "multivariate",
                                years_col = "OS_years",
                                status_col = "OS_status")
    
    plotForest(readRDS(file.path(results_dir, glue::glue("coxph_add_OS_mb_{abbrev}_resection_metastasis_plp.RDS"))))
  
    ggsave(file.path(plot_dir, glue::glue("forest_OS_mb_{abbrev}_resection_subtype_metastasis.pdf")),
           width = 8, height = 3, units = "in")
    
  }
  
}
```
Plot KM survival curves for MB SHH, stratified by infant vs child/adult type 

```{r}
kap_shh_type_os <- survival_analysis(
  metadata  = mb_hist[!is.na(mb_hist$shh_type),],
  ind_var = "shh_type",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal", 
  days_col = "OS_days",
  status_col = "OS_status"
)

kap_shh_type_efs <- survival_analysis(
  metadata  = mb_hist[!is.na(mb_hist$shh_type),],
  ind_var = "shh_type",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal", 
  days_col = "EFS_days",
  status_col = "EFS_status"
)

km_shh_type_plot <- plotKM(model = list(kap_shh_type_os,
                                      kap_shh_type_efs),
                  variable = "shh_type",
                  combined = T, 
                  title = "MB, SHH")
km_shh_type_plot

ggsave(file.path(plot_dir, "km_OS_mb_shh_subtype.pdf"),
       width = 9, height = 5)
```
Plot KM survival curves for MB, Group 4 cohort by methylation subtype

```{r}
kap_g4_os <- survival_analysis(
  metadata  = mb_hist[mb_hist$molecular_subtype == "MB, Group4" & grepl("G34", mb_hist$dkfz_v12_methylation_subclass),],
  ind_var = "dkfz_v12_methylation_subclass",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal", 
  days_col = "OS_days",
  status_col = "OS_status"
)

kap_g4_efs <- survival_analysis(
  metadata  = mb_hist[mb_hist$molecular_subtype == "MB, Group4" & grepl("G34", mb_hist$dkfz_v12_methylation_subclass),],
  ind_var = "dkfz_v12_methylation_subclass",
  test = "kap.meier",
  metadata_sample_col = "Kids_First_Biospecimen_ID_normal", 
  days_col = "EFS_days",
  status_col = "EFS_status"
)

km_g4_plot <- plotKM(model = list(kap_g4_os,
                                      kap_g4_efs),
                  variable = "dkfz_v12_methylation_subclass",
                  combined = T, 
                  title = "MB, Group4")
km_g4_plot

ggsave(file.path(plot_dir, "km_mb_g4_subtype.pdf"),
       width = 8, height = 5.5)
```
Plot MB Group 4 EFS coxph model including methylation subclass as a covariate

```{r}
add_mb_g4_efs_model <- fit_save_model(mb_hist[mb_hist$molecular_subtype == "MB, Group4" & grepl("G34", mb_hist$dkfz_v12_methylation_subclass),],
                                "extent_of_tumor_resection+dkfz_v12_methylation_subclass+cpg_plp",
                                file.path(results_dir, "coxph_add_EFS_mb_g4_methyl_subtype.RDS"),
                                "multivariate",
                                years_col = "EFS_years",
                                status_col = "EFS_status")

plotForest(readRDS(file.path(results_dir, "coxph_add_EFS_mb_g4_methyl_subtype.RDS")))

ggsave(file.path(plot_dir, "forest_EFS_mb_g4_resection_methyl_subtype.pdf"),
       width = 7, height = 3, units = "in")
```
Plot MB Group 3 EFS coxph model including methylation subtype as covariate

```{r}
add_mb_g3_efs_model <- fit_save_model(mb_hist[mb_hist$molecular_subtype == "MB, Group3" & grepl("G34", mb_hist$dkfz_v12_methylation_subclass),],
                                "extent_of_tumor_resection+dkfz_v12_methylation_subclass+cpg_plp",
                                file.path(results_dir, "coxph_add_EFS_mb_g3_resection_methyl_subtype.RDS"),
                                "multivariate",
                                years_col = "EFS_years",
                                status_col = "EFS_status")

plotForest(readRDS(file.path(results_dir, "coxph_add_EFS_mb_g3_resection_methyl_subtype.RDS")))

ggsave(file.path(plot_dir, "forest_EFS_mb_g3_resection_methyl_subtype.pdf"),
       width = 7, height = 3, units = "in")
```
Plot MB Group 3 OS coxph model including methylation subtype as covariate

```{r}
add_mb_g3_efs_model <- fit_save_model(mb_hist[mb_hist$molecular_subtype == "MB, Group3" & grepl("G34", mb_hist$dkfz_v12_methylation_subclass),],
                                "extent_of_tumor_resection+dkfz_v12_methylation_subclass+cpg_plp",
                                file.path(results_dir, "coxph_add_OS_mb_g3_resection_methyl_subtype.RDS"),
                                "multivariate",
                                years_col = "OS_years",
                                status_col = "OS_status")

plotForest(readRDS(file.path(results_dir, "coxph_add_OS_mb_g3_resection_methyl_subtype.RDS")))

ggsave(file.path(plot_dir, "forest_OS_mb_g3_resection_methyl_subtype.pdf"),
       width = 7, height = 3, units = "in")
```

```{r}
sessionInfo()
```