---
title: 'Prevalence of germline P-LP variants in predisposition-syndrome-associated genes'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2023"
---

Load packages and set directories
```{r load libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(tidyverse)
  library(stringr)
})

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "predisposition-variants")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
```

Set paths for: 
- histologies file, 
- germline plp variant file
- predisposition syndromes files

```{r set paths}
histologies_file <- file.path(root_dir, 
                           "analyses", 
                           "collapse-tumor-histologies", 
                           "results",
                           "germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

plp_file <- file.path(data_dir, 
                      "pbta-merged-plp-variants-autogvp-abridged.tsv")

plp_sv_file <- file.path(data_dir, 
                         "pbta_germline_svs.tsv")

predispositions_file <- file.path(input_dir, "predisposition-syndromes.tsv")
```

Read in histologies, germline PLP, and predispositions file

```{r read files}
hist <- read_tsv(histologies_file,show_col_types = FALSE) 

plp <- read_tsv(plp_file, show_col_types = FALSE)

plp_sv <- read_tsv(plp_sv_file, show_col_types = FALSE)

predispositions <- read_tsv(predispositions_file) %>%
  dplyr::rename(gene_symbol_vep = Hugo_Symbol)
```

See which cancer predispositions are currently reported in histologies file: 
```{r get predispositions}
table(hist$cancer_predispositions)
```

Subset PLP file to include only variants in CPGs associated with syndromes in `predispositions`, and include `cancer_predispositions` column to indicate whether syndrome is currently reported.

```{r }
predisposition_plp_bs <- plp %>%
  filter(gene_symbol_vep %in% predispositions$gene_symbol_vep) %>%
  pull(Kids_First_Biospecimen_ID_normal)

plp_predisposition_syndromes <- hist %>%
  filter(Kids_First_Biospecimen_ID_normal %in% predisposition_plp_bs) %>%
  left_join(plp[,c("Kids_First_Biospecimen_ID_normal", "chr", "start",
                   "ref", "alt", "gene_symbol_vep", "autogvp_call", 
                   "autogvp_call_reason", "HGVSc", "HGVSp",
                   "clinvar_variant_id")]) %>%
  left_join(predispositions) %>%
  filter(gene_symbol_vep %in% predispositions$gene_symbol_vep) %>%
  dplyr::rename("associated_syndrome" = `Associated Syndrome`,
                "clinically_reported_syndrome" = cancer_predispositions) %>%
  dplyr::mutate(incidental_finding = case_when(
    !str_detect(clinically_reported_syndrome, associated_syndrome) ~ "Yes",
    TRUE ~ "No"
  )) %>%
  dplyr::select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID_normal, 
         sample_id_normal, sample_id_tumor,
         plot_group, molecular_subtype, 
         chr, start, alt, ref,
         gene_symbol_vep, autogvp_call, autogvp_call_reason, HGVSc, HGVSp,
         clinvar_variant_id,
         associated_syndrome, clinically_reported_syndrome, incidental_finding) %>%
  dplyr::mutate(clinvar_variant_link = glue::glue("https://www.ncbi.nlm.nih.gov/clinvar/variation/{clinvar_variant_id}/"))
```

Save output

```{r}
write_tsv(plp_predisposition_syndromes,
          file.path(results_dir, "incidental-findings-predisposition-variants.tsv"))
```


Generate same file for germline P/LP SVs: 

```{r}
predisposition_plp_sv_bs <- plp_sv %>%
  filter(Hugo_Symbol_cpg %in% predispositions$gene_symbol_vep) %>%
  pull(Kids_First_Biospecimen_ID_normal)

plp_sv_predisposition_syndromes <- hist %>%
  filter(Kids_First_Biospecimen_ID_normal %in% predisposition_plp_sv_bs) %>%
  left_join(plp_sv) %>%
  dplyr::rename(gene_symbol_vep = Hugo_Symbol_cpg) %>%
  left_join(predispositions) %>%
  filter(gene_symbol_vep %in% predispositions$gene_symbol_vep) %>%
  dplyr::rename("associated_syndrome" = `Associated Syndrome`,
                "clinically_reported_syndrome" = cancer_predispositions) %>%
  dplyr::mutate(incidental_finding = case_when(
    !str_detect(clinically_reported_syndrome, associated_syndrome) ~ "Yes",
    TRUE ~ "No"
  )) %>%
  dplyr::select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID_normal, 
         Kids_First_Biospecimen_ID_tumor,
         sample_id_normal, sample_id_tumor,
         plot_group, molecular_subtype, 
         Chromosome, Start, End, Length,
         Type, Classification, 
         gene_symbol_vep, 
         associated_syndrome, clinically_reported_syndrome, incidental_finding)
```

Save output

```{r}
write_tsv(plp_sv_predisposition_syndromes,
          file.path(results_dir, "incidental-findings-predisposition-structural-variants.tsv"))
```
