---
title: 'Plot predisposition syndrome-associated gene P/LP prevalence across germline cohort'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2023"
---

Load packages and set directories
```{r load libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(tidyverse)
  library(ggplot2)
  library(ggsci)
})

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "predisposition-variants")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results") 
plots_dir <- file.path(analysis_dir, "plots")

source(file.path(root_dir, "figures", "theme.R"))
```

Set paths for: 
- histologies file, 
- germline plp variant file
```{r set paths}
histologies_file <- file.path(root_dir, "analyses", 
                              "collapse-tumor-histologies", "results",
                              "germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

plp_file <- file.path(data_dir, 
                      "pbta-merged-plp-variants-autogvp-abridged.tsv")

plp_lowVAF_file <- file.path(data_dir, 
                             "pbta-merged-plp-variants-autogvp-abridged-lowVAF-cpgs.tsv")

plp_sv_file <- file.path(root_dir, "analyses", 
                         "germline-sv", "input", 
                         "pbta_germline_svs.tsv")

predispositions_file <- file.path(input_dir, "predisposition-syndromes.tsv")

predispositions_path_review_file <- file.path(input_dir, "predispositions-path-review.tsv")
```

Load files
```{r read files}
hist <- read_tsv(histologies_file, guess_max = 10000, 
                 show_col_types = FALSE) %>%
  dplyr::mutate(cancer_predispositions = case_when(
    grepl("Constitutional|Lynch", cancer_predispositions) ~ "CMMRD/Lynch Syndrome",
    TRUE ~ cancer_predispositions
  ))

plp_lowVAF <- read_tsv(plp_lowVAF_file, show_col_types = FALSE)

plp <- read_tsv(plp_file, show_col_types = FALSE) %>%
  bind_rows(plp_lowVAF)

plp_sv <- read_tsv(plp_sv_file, show_col_types = FALSE)

syndromes <- read_tsv(predispositions_file) %>%
  # Abbreviate syndromes for plotting
  dplyr::mutate(`Associated Syndrome` = case_when(
    `Associated Syndrome` == "Familial Adenomatous Polyposis" ~ "FAP",
    `Associated Syndrome` == "Rhabdoid Tumor Predisposition syndrome" ~ "RTPS",
    `Associated Syndrome` == "ELP1 Medulloblastoma Syndrome" ~ "ELP1 MB syndrome",
    `Associated Syndrome` == "Melanoma Astrocytoma Syndrome" ~ "MAS",
    `Associated Syndrome` == "BAP1 Tumor Predisposition Syndrome" ~ "BAP1-TPS",
    TRUE ~ `Associated Syndrome`
  )) %>%
  dplyr::rename(gene_symbol_vep = Hugo_Symbol)

syndrome_genes <- syndromes$gene_symbol_vep

predispositions_path_review <- read_tsv(predispositions_path_review_file) %>%
  dplyr::rename(gene_symbol_vep = Hugo_Symbol)
```

## Assess frequency of associated P/LP variants in patients with reported cancer predisposition syndromes

Define cancer predispositions and associated genes
```{r}
predisposition_genes <- c("CMMRD/Lynch Syndrome" = "PMS2|MLH1|MSH2|MSH6",
                          "Familial Adenomatous Polyposis" = "APC",
                          "Gorlin Syndrome" = "PTCH1|SUFU",
                          "Li-Fraumeni syndrome" = "TP53",
                          "NF-1" = "NF1", "NF-2" = "NF2",
                          "Schwannomatosis" = "LZTR1",
                          "Tuberous Sclerosis" = "TSC1|TSC2",
                          "Von Hippel-Lindau" = "VHL")

predispositions <- names(predisposition_genes)
```

Loop through cancer predispositions and create a data frame indicating presence/absence of associated gene germline P/LP variant
```{r}
predisposition_list <- list()

suppressWarnings(
  for (predisposition in predispositions) {
    
  genes <- unlist(strsplit(predisposition_genes[predisposition], "[|]"))
  
  # get pts confirmed to have syndrome through path report review 
  
  predisposition_list[[predisposition]] <- hist %>%
    filter(grepl(predisposition, cancer_predispositions)) %>%
    dplyr::select(Kids_First_Participant_ID, Kids_First_Biospecimen_ID_normal, sample_id_normal, sample_id_tumor, plot_group, molecular_subtype) %>%
    dplyr::mutate(predisposition = predisposition) %>%
    dplyr::mutate(plp_variant = case_when(
      Kids_First_Biospecimen_ID_normal %in% c(plp$Kids_First_Biospecimen_ID_normal[plp$gene_symbol_vep %in% genes],
                                              plp_sv$Kids_First_Biospecimen_ID_normal[plp_sv$Hugo_Symbol_cpg %in% genes])~ "Yes",
      TRUE ~ "No"
    ))
  }
)

# cat results
predisposition_df <- Reduce(rbind, predisposition_list)

```

Filter for patients that do not show P-LP variant associated with predisposition syndrome, and write to output
```{r}
predisposition_df %>%
  dplyr::filter(plp_variant == "No") %>%
  write_tsv(file.path(results_dir, "predisposition-patients-no-plp-variants.tsv"))
```

Create bar plot of patients with/without P/LP germline variants in gene associated with each cancer predisposition syndrome
```{r}
labels <- c("Von Hippel-Lindau (VHL)",
            "Schwannomatosis (LZTR1)",
            "Familial Adenomatous Polyposis (APC)",
            "Gorlin Syndrome (PTCH1,SUFU)",
            "CMMRD/Lynch Syndrome (PMS2, MLH1, MSH2, MSH6)",
            "NF-2 (NF2)",
            "Li-Fraumeni Syndrome (TP53)",
            "Tuberous Sclerosis (TSC1/2)",
            "NF-1 (NF1)")

predisposition_df %>%
  count(predisposition, plp_variant) %>%
  dplyr::mutate(predisposition = fct_relevel(predisposition,
                                      c("Von Hippel-Lindau", "Schwannomatosis",
                                        "Familial Adenomatous Polyposis",
                                        "Gorlin Syndrome", 
                                        "CMMRD/Lynch Syndrome",
                                        "NF-2", "Li-Fraumeni syndrome",
                                        "Tuberous Sclerosis", "NF-1"))) %>%
  dplyr::mutate(plp_variant = fct_relevel(plp_variant,
                                      c("Yes", "No"))) %>%
  ggplot(aes(x = n, y = predisposition, fill = plp_variant)) + 
  geom_bar(stat = "identity", 
           colour = "black") +
  labs(x = "No. patients",
       y = "",
       fill = "P/LP variant") +
  scale_y_discrete(labels=labels) +
  scale_fill_npg() +
  theme_Publication() +
  theme(legend.position = "top") +
  guides(fill=guide_legend(ncol=2)) +
  xlim(c(0,30))

ggsave(file.path(plots_dir, "gene-plp-prevalence-by-syndrome.pdf"),
       width = 7, height = 4)
```

## Plot frequency of incidental findings in cohort

Load incidental findings summary, and change incidental findings to "No" when predisposition is recorded in path review 
```{r}
incidental_df <- read_tsv(file.path(results_dir, "incidental-findings-predisposition-variants.tsv"))

incidental_df <- incidental_df %>%
  left_join(predispositions_path_review[,c("Kids_First_Biospecimen_ID_normal", "gene_symbol_vep", "cancer_predispositions_path")]) %>%
  dplyr::rename("path_report_findings" = cancer_predispositions_path) %>%
  dplyr::mutate(path_report_findings = case_when(
    !is.na(path_report_findings) ~ glue::glue("{path_report_findings} confirmed"),
    is.na(path_report_findings) ~ ""
  ),
  incidental_finding = case_when(
    incidental_finding == "No" | path_report_findings != "" ~ "No",
    TRUE ~ "Yes"
  )) %>%
    # retain only unique pt-gene pairs (remove cases of multi-hits to gene in same pt)
  distinct(Kids_First_Participant_ID, gene_symbol_vep, .keep_all = T)

# Save update incidental findings
write_tsv(incidental_df,
          file.path(results_dir, "incidental-findings-predisposition-variants-path-review.tsv"))

```

Load incidental findings summary, and change incidental findings to "No" when predisposition is recorded in path review 
```{r}
incidental_sv_df <- read_tsv(file.path(results_dir, "incidental-findings-predisposition-structural-variants.tsv"))

incidental_sv_df <- incidental_sv_df %>%
  left_join(predispositions_path_review[,c("Kids_First_Biospecimen_ID_normal", "gene_symbol_vep", "cancer_predispositions_path")]) %>%
  dplyr::rename("path_report_findings" = cancer_predispositions_path) %>%
  dplyr::mutate(path_report_findings = case_when(
    !is.na(path_report_findings) ~ glue::glue("{path_report_findings} confirmed"),
    is.na(path_report_findings) ~ ""
  ),
  incidental_finding = case_when(
    incidental_finding == "No" | path_report_findings != "" ~ "No",
    TRUE ~ "Yes"
  )) %>%
    # retain only unique pt-gene pairs (remove cases of multi-hits to gene in same pt)
  distinct(Kids_First_Participant_ID, gene_symbol_vep, .keep_all = T)

# Save update incidental findings
write_tsv(incidental_sv_df,
          file.path(results_dir, "incidental-findings-predisposition-structural-variants-path-review.tsv"))

```

How many of these findings are incidental? 
```{r}
table(incidental_df$incidental_finding)

table(incidental_sv_df$incidental_finding)

```

Plot predisposition gene variant count, and distinguish between patients who have predisposition reported vs. those who do not
```{r}
# Calculate predisposition gene counts for ordering plot
total_cts <- incidental_df %>%
  dplyr::select(Kids_First_Biospecimen_ID_normal, gene_symbol_vep) %>%
  bind_rows(incidental_sv_df %>% dplyr::select(Kids_First_Biospecimen_ID_normal, gene_symbol_vep)) %>%
  distinct(Kids_First_Biospecimen_ID_normal, gene_symbol_vep) %>%
  count(gene_symbol_vep) %>%
  arrange(n)

# Define plot labels
labels <- glue::glue("{syndrome_genes} ({syndromes$`Associated Syndrome`})")
names(labels) <- syndrome_genes

incidental_df %>%
  dplyr::select(Kids_First_Biospecimen_ID_normal, gene_symbol_vep, incidental_finding) %>%
  bind_rows(incidental_sv_df %>% dplyr::select(Kids_First_Biospecimen_ID_normal, 
                                               gene_symbol_vep, incidental_finding)) %>%
  distinct(Kids_First_Biospecimen_ID_normal, gene_symbol_vep, .keep_all = T) %>%
  dplyr::mutate(predisposition = case_when(
    incidental_finding == "Yes" ~ "No",
    incidental_finding == "No" ~ "Yes"
  )) %>%
  count(gene_symbol_vep, predisposition) %>%
  dplyr::mutate(gene_symbol_vep = fct_relevel(gene_symbol_vep,
                                      total_cts$gene_symbol_vep)) %>%
  dplyr::mutate(predisposition = fct_relevel(predisposition,
                                      c("Yes", "No"))) %>%
  ggplot(aes(x = n, y = gene_symbol_vep, fill = predisposition)) + 
  geom_bar(stat = "identity", 
           colour = "black") +
  labs(x = "No. patients",
       y = "",
       fill = "Predisposition clinically-reported") +
  xlim(c(0,25)) +
  scale_y_discrete(labels=labels) +
  scale_fill_npg() +
  theme_Publication() +
  theme(legend.position = "top") +
  guides(fill=guide_legend(ncol=3))

ggsave(file.path(plots_dir, "syndrome-prevalence-by-gene.pdf"),
       width = 7, height = 6)
```

```{r}
sessionInfo()

```

