---
title: "Filter P-LP variants"
author: Ryan Corbett
output: html_document
date: "2024"
---

This script queries P-LP variants for overlap with exome capture regions, and removes those that fall outside regions to be used for gnomAD and PMBB enrichment analyses


# Load libraries
```{r load libraries, set up directories}
suppressPackageStartupMessages({
  library(tidyverse)
  library(GenomicRanges)
  library(IRanges)
})

# Set up directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "bed-intersect")
results_dir <- file.path(analysis_dir, "results")

# If the directory does not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}
```

Set file paths

```{r}
plp_file <- file.path(data_dir, 
                      "pbta-merged-plp-variants-autogvp-abridged.tsv")

hist_file <- file.path(data_dir,
                       "histologies.tsv")

hist_cohort_file <- file.path(root_dir, "analyses",
                             "collapse-tumor-histologies",
                             "results", "germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

wxs_region_file <- file.path(root_dir, "analyses",
                             "bed-intersect", "results",
                             "pmbb_pbta_wxs_intersect.bed")
```

Wrangle data

```{r}
hist <- read_tsv(hist_file)

hist_cohort <- read_tsv(hist_cohort_file)

plp <- read_tsv(plp_file) %>%
  dplyr::filter(Kids_First_Biospecimen_ID_normal %in% hist_cohort$Kids_First_Biospecimen_ID_normal)
```

Obtain WXS ids to use for filtering later

```{r}
wxs_ids <- hist %>%
  dplyr::filter(experimental_strategy == "WXS") %>%
  pull(Kids_First_Biospecimen_ID)
```

Create GRanges object from P-LP file, using `start` value as start and end

```{r}
plp_coords <- plp %>%
  dplyr::mutate(end = start) %>%
  dplyr::select(chr, start, end, gene_symbol_vep)

plp_granges <- makeGRangesFromDataFrame(plp_coords,
                                        keep.extra.columns = TRUE)
```

Load exome capture intersection bed file, and convert to GRanges object

```{r}
exome_bed <- read_tsv(wxs_region_file,
                      col_names = c("chr", "start", "end"))

exome_granges <- makeGRangesFromDataFrame(exome_bed)
```

Find overlaps between P-LP variant coordinates and exome capture regions

```{r}
plp_exome_intersect <- findOverlaps(plp_granges, exome_granges)

intersect_df <- data.frame(plp_exome_intersect)

head(intersect_df)
```

The `queryHits` column of `intersect_df` lists the row names of P-LP variants in `plp` that overlap with exome capture regions.
We can extract these row numbers and get corresponding variant ids to use for filtering

```{r}
to_retain <- intersect_df$queryHits

variants_to_rm <- plp[-to_retain,] %>%
  pull(variant_id)
```

We will generate two filtered data frames:
1) P-LP variants excluding those that do not overlap exome capture regions (to use for PMBB enrichment analyses)
2) P-LP variants excluding only WXS-derived variants that do not overlap exome capture regions (for gnomAD enrichment analyses)

```{r}
plp_all_exome_filtered <- plp %>%
  dplyr::filter(!variant_id %in% variants_to_rm) %>%
  write_tsv(file.path(results_dir, "pbta-merged-plp-variants-autogvp-abridged-all-exome-filtered.tsv"))

plp_wxs_exome_filtered <- plp %>%
  dplyr::filter(!variant_id %in% variants_to_rm | !Kids_First_Biospecimen_ID_normal %in% wxs_ids) %>%
  write_tsv(file.path(results_dir, "pbta-merged-plp-variants-autogvp-abridged-wxs-exome-filtered.tsv"))
```

```{r}
sessionInfo()
```