---
title: "Filter P-LP variants"
author: Ryan Corbett
output: html_document
date: "2024"
---

This script queries P-LP variants for overlap with exome capture regions, and removes those that fall outside regions to be used for gnomAD and PMBB enrichment analyses


# Load libraries
```{r load libraries, set up directories}
suppressPackageStartupMessages({
  library(tidyverse)
  library(GenomicRanges)
  library(IRanges)
})

# Set up directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "bed-intersect")
results_dir <- file.path(analysis_dir, "results")

# If the directory does not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}
```

Set file paths

```{r}
plp_file <- file.path(data_dir, 
                      "pbta-merged-plp-variants-autogvp-abridged.tsv")

hist_file <- file.path(data_dir,
                       "histologies.tsv")

hist_cohort_file <- file.path(root_dir, "analyses",
                             "collapse-tumor-histologies",
                             "results", "germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

wxs_region_20bp_file <- file.path(results_dir,
                             "pmbb_pbta_wxs_intersect.20bp_padded.bed")

wxs_region_100bp_file <- file.path(results_dir,
                             "pmbb_pbta_wxs_intersect.100bp_padded.bed")

pbta_region_file <- file.path(results_dir, 
                              "pbta_wxs_intersect.bed")
```

Wrangle data

```{r}
hist <- read_tsv(hist_file)

hist_cohort <- read_tsv(hist_cohort_file)

plp <- read_tsv(plp_file) %>%
  dplyr::filter(Kids_First_Biospecimen_ID_normal %in% hist_cohort$Kids_First_Biospecimen_ID_normal)
```

Obtain WXS ids to use for filtering later

```{r}
wxs_ids <- hist %>%
  dplyr::filter(experimental_strategy == "WXS") %>%
  pull(Kids_First_Biospecimen_ID)
```

Create GRanges object from P-LP file, using `start` value as start and end

```{r}
plp_coords <- plp %>%
  dplyr::mutate(end = start) %>%
  dplyr::select(chr, start, end, gene_symbol_vep)

plp_granges <- makeGRangesFromDataFrame(plp_coords,
                                        keep.extra.columns = TRUE)
```

Load exome capture intersection bed file, and convert to GRanges object

```{r}
exome_20bp_bed <- read_tsv(wxs_region_20bp_file,
                      col_names = c("chr", "start", "end"))

exome_20bp_granges <- makeGRangesFromDataFrame(exome_20bp_bed)

exome_100bp_bed <- read_tsv(wxs_region_100bp_file,
                      col_names = c("chr", "start", "end"))

exome_100bp_granges <- makeGRangesFromDataFrame(exome_100bp_bed)
```

Find overlaps between P-LP variant coordinates and exome capture regions

```{r}
exome_granges_list <- list(exome_20bp_granges, 
                           exome_100bp_granges)

padding <- c("20bp_padded", "100bp_padded")

for (i in 1:length(exome_granges_list)){
  
  plp_exome_intersect <- findOverlaps(plp_granges, exome_granges_list[[i]])

  pmbb_intersect_df <- data.frame(plp_exome_intersect)
  
  # The `queryHits` column of `pmbb_intersect_df` lists the row names of P-LP variants in `plp` that overlap with exome capture regions.
  # We can extract these row numbers and get corresponding variant ids to use for filtering
  
  to_retain_pmbb <- pmbb_intersect_df$queryHits

  variants_to_rm_pmbb <- plp[-to_retain_pmbb,] %>%
    pull(variant_id)
  
  #   We will generate two filtered data frames:
  # 1) P-LP variants excluding those that do not overlap exome capture regions (to use for PMBB enrichment analyses)
  # 2) P-LP variants excluding only WXS-derived variants that do not overlap exome capture regions (for gnomAD enrichment analyses)
  
  plp_all_pmbb_exome_filtered <- plp %>%
    dplyr::filter(!variant_id %in% variants_to_rm_pmbb) %>%
    write_tsv(file.path(results_dir, 
                        glue::glue("pbta-merged-plp-variants-autogvp-abridged-all-exome-filtered-{padding[i]}.tsv")))

  plp_wxs_pmbb_exome_filtered <- plp %>%
    dplyr::filter(!variant_id %in% variants_to_rm_pmbb | !Kids_First_Biospecimen_ID_normal %in% wxs_ids) %>%
    write_tsv(file.path(results_dir, 
                        glue::glue("pbta-merged-plp-variants-autogvp-abridged-wxs-exome-filtered-{padding[i]}.tsv")))
  
  # write variants that were excluded to output
  
  plp %>%
    dplyr::filter(variant_id %in% variants_to_rm_pmbb) %>%
    write_tsv(file.path(results_dir, 
                        glue::glue("pbta-merged-plp-variants-autogvp-abridged-all-exome-excluded-{padding[i]}.tsv")))

  plp %>%
    dplyr::filter(variant_id %in% variants_to_rm_pmbb & Kids_First_Biospecimen_ID_normal %in% wxs_ids) %>%
    write_tsv(file.path(results_dir, 
                        glue::glue("pbta-merged-plp-variants-autogvp-abridged-wxs-exome-excluded-{padding[i]}.tsv")))
  
}
```

Load PBTA exome capture region intersect bed, convert to GRanges, and get overlap with P-LP variants

```{r}
pbta_exome_bed <- read_tsv(pbta_region_file,
                      col_names = c("chr", "start", "end"))

pbta_exome_granges <- makeGRangesFromDataFrame(pbta_exome_bed)

plp_exome_intersect <- findOverlaps(plp_granges, pbta_exome_granges)

pbta_intersect_df <- data.frame(plp_exome_intersect)
```

Get list of P-LP variants to exclude

```{r}
to_retain_gnomad <- pbta_intersect_df$queryHits

variants_to_rm_gnomad <- plp[-to_retain_gnomad,] %>%
  pull(variant_id)
```

Create two output for gnomAD enrichment analyses:
1) P-LP variants in WGS samples ONLY (*no-wxs.tsv)
2) P-LP variants in PBTA exome capture regions (*pbta-exome-filtered.tsv)

```{r}
plp_wxs_filtered <- plp %>%
  dplyr::filter(!Kids_First_Biospecimen_ID_normal %in% wxs_ids) %>%
  write_tsv(file.path(results_dir, 
                        glue::glue("pbta-merged-plp-variants-autogvp-abridged-no-wxs.tsv")))

plp_pbta_exome_filtered <- plp %>%
  dplyr::filter(!variant_id %in% variants_to_rm_gnomad) %>%
  write_tsv(file.path(results_dir, 
                        glue::glue("pbta-merged-plp-variants-autogvp-abridged-pbta-exome-filtered.tsv")))

# write all variants outside of exome capture regions 
plp %>%
  dplyr::filter(variant_id %in% variants_to_rm_gnomad) %>%
  write_tsv(file.path(results_dir, 
                        glue::glue("pbta-merged-plp-variants-autogvp-abridged-pbta-exome-excluded.tsv")))
```


```{r}
sessionInfo()
```