---
title: 'CPG PLP enrichment analyses'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "February 2023"
---
  
Load packages and set directories

```{r load libraries}
library(data.table)
library(tidyverse)

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))
setwd(root_dir)

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "cpg-enrichment")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
```

Set histologies, germline plp, cpg, and control cohort cpg count file paths

```{r set file paths}
cbtn_histologies_file <- file.path(root_dir,"analyses", "collapse-tumor-histologies", "results", "germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

plp_file <- file.path(data_dir, "v3", "pbta_germline_plp_calls.tsv")

cpg_file <- file.path(root_dir, "analyses", "oncokb-annotation", "input", "cpg.txt")

gnomad_file <- file.path(input_dir, "PBTA-CPG-PLP-enrichment-gnomAD.tsv")

pmbb_file <- file.path(input_dir, "PBTA-CPG-PLP-enrichment-PMBB.tsv")
```

Read in plp file and filter for CPGs and 838 samples in hist

```{r filter plp}
cpgs <- read_lines(cpg_file)

hist <- read_tsv(cbtn_histologies_file)

plp <- read_tsv(plp_file) %>%
  filter(Hugo_Symbol %in% cpgs & Kids_First_Biospecimen_ID %in% hist$Kids_First_Biospecimen_ID_normal)
```

Read in gnomad cohort cpg plp frequencies and run Fisher's exact tests for each CPG  

```{r gnomad enrichment}

# read in gnomad freqs

gnomad_cts <- read_tsv(gnomad_file) %>%
  filter(Hugo_symbol %in% cpgs)

# merge cohort counts and gnomad freqs for each cpg 

cpg_enr_gnomad <- plp %>%
  count(Hugo_Symbol) %>%
  mutate(n_without = 838-n) %>%
  dplyr::rename("Hugo_symbol" = Hugo_Symbol) %>%
  left_join(gnomad_cts[,c("Hugo_symbol", "count_with_plp_gnomAD", "count_without_plp_gnomAD")], by = "Hugo_symbol") %>%
  distinct(Hugo_symbol, .keep_all = TRUE) %>%
  mutate(oddsRatio = 0, pvalue = 0, CI_lower = 0, CI_upper = 0)

# run Fisher's exact tests for each cpg 

for (i in 1:nrow(cpg_enr_gnomad)){
  if(!is.na(cpg_enr_gnomad$count_with_plp_gnomAD[i])) {
    mat <- matrix(c(cpg_enr_gnomad$n[i], cpg_enr_gnomad$n_without[i],
                    cpg_enr_gnomad$count_with_plp_gnomAD[i], cpg_enr_gnomad$count_without_plp_gnomAD[i]),
                  2,2)
    ftest <- fisher.test(mat)
    cpg_enr_gnomad$oddsRatio[i] <- ftest$estimate
    cpg_enr_gnomad$pvalue[i] <- ftest$p.value
    cpg_enr_gnomad$CI_lower[i] <- ftest$conf.int[1]
    cpg_enr_gnomad$CI_upper[i] <- ftest$conf.int[2]
  }
}

# calculate FDRs

cpg_enr_gnomad <- cpg_enr_gnomad %>%
  mutate(adjusted_p = p.adjust(pvalue, method = "bonferroni"))

# save result

write_tsv(cpg_enr_gnomad, 
          file.path(results_dir, "PBTA-cpg-plp-enrichment-gnomAD.tsv"))
```

Read in PMBB cohort cpg plp frequencies and run Fisher's exact tests for each CPG  

```{r pmbb enrichment}

# read in PMBB freqs

pmbb_cts <- read_tsv(pmbb_file)

# merge cohort counts and PMBB freqs for each cpg 

cpg_enr_pmbb <- plp %>%
  count(Hugo_Symbol) %>%
  mutate(n_without = 838-n) %>%
  dplyr::rename("Hugo_symbol" = Hugo_Symbol) %>%
  left_join(pmbb_cts[,c("Hugo_symbol", "count_with_plp_PMBB_noTumor", "count_without_plp_PMBB_noTumor")], by = "Hugo_symbol") %>%
  distinct(Hugo_symbol, .keep_all = TRUE) %>%
  mutate(oddsRatio = 0, pvalue = 0, CI_lower = 0, CI_upper = 0)

# run Fisher's exact tests for each cpg 

for (i in 1:nrow(cpg_enr_pmbb)){
  if(!is.na(cpg_enr_pmbb$count_with_plp_PMBB_noTumor[i])) {
    mat <- matrix(c(cpg_enr_pmbb$n[i], cpg_enr_pmbb$n_without[i],
                    cpg_enr_pmbb$count_with_plp_PMBB_noTumor[i], cpg_enr_pmbb$count_without_plp_PMBB_noTumor[i]),
                  2,2)
    ftest <- fisher.test(mat)
    cpg_enr_pmbb$oddsRatio[i] <- ftest$estimate
    cpg_enr_pmbb$pvalue[i] <- ftest$p.value
    cpg_enr_pmbb$CI_lower[i] <- ftest$conf.int[1]
    cpg_enr_pmbb$CI_upper[i] <- ftest$conf.int[2]
  }
}

# calculate FDRs

cpg_enr_pmbb <- cpg_enr_pmbb %>%
  mutate(adjusted_p = p.adjust(pvalue, method = "bonferroni"))

# save result

write_tsv(cpg_enr_pmbb, 
          file.path(results_dir, "PBTA-cpg-plp-enrichment-PMBB.tsv"))
```

Calculate CPG PLP enrichment in CBTN cohort relative to gnomAD by gene and histology group  

```{r gnomad enrichment by hist}

# calculate number of patients per plot group

hist_cts <- hist %>%
  count(plot_group) %>%
  rename("hist_n" = n)

# obtain counts of plp variants by plot group and cpg

hist_cpg_enr_gnomad <- plp %>%
  dplyr::rename("Kids_First_Biospecimen_ID_normal" = Kids_First_Biospecimen_ID) %>%
  left_join(hist[,c("Kids_First_Biospecimen_ID_normal", "plot_group")], by = "Kids_First_Biospecimen_ID_normal") %>%
  count(plot_group, Hugo_Symbol) %>%
  left_join(hist_cts, by = "plot_group") %>%
  mutate(n_without = hist_n - n) %>%
  dplyr::rename("Hugo_symbol" = Hugo_Symbol) %>%
  left_join(gnomad_cts[,c("Hugo_symbol", "count_with_plp_gnomAD", "count_without_plp_gnomAD")], by = "Hugo_symbol") %>%
  distinct(plot_group, Hugo_symbol, .keep_all = TRUE) %>%
  mutate(oddsRatio = 0, pvalue = 0, CI_lower = 0, CI_upper = 0)

# run Fisher's exact test for each cpg and plot group

for (i in 1:nrow(hist_cpg_enr_gnomad)){
  if(!is.na(hist_cpg_enr_gnomad$count_with_plp_gnomAD[i])) {
    mat <- matrix(c(hist_cpg_enr_gnomad$n[i], hist_cpg_enr_gnomad$n_without[i],
                    hist_cpg_enr_gnomad$count_with_plp_gnomAD[i], hist_cpg_enr_gnomad$count_without_plp_gnomAD[i]),
                  2,2)
    ftest <- fisher.test(mat)
    hist_cpg_enr_gnomad$oddsRatio[i] <- ftest$estimate
    hist_cpg_enr_gnomad$pvalue[i] <- ftest$p.value
    hist_cpg_enr_gnomad$CI_lower[i] <- ftest$conf.int[1]
    hist_cpg_enr_gnomad$CI_upper[i] <- ftest$conf.int[2]
  }
}

# calculate FDR 

hist_cpg_enr_gnomad <- hist_cpg_enr_gnomad %>%
  mutate(adjusted_p = p.adjust(hist_cpg_enr_gnomad$pvalue, method = "bonferroni"))

# save output 

write_tsv(hist_cpg_enr_gnomad, 
          file.path(results_dir, "PBTA-cpg-plp-enrichment-by-plotGroup-gnomAD.tsv"))
```

Calculate CPG PLP enrichment in CBTN cohort relative to PMBB by gene and histology group  

```{r pmbb enrichment by hist}

# obtain counts of plp variants by plot group and cpg and merge with PMBB

hist_cpg_enr_pmbb <- plp %>%
  dplyr::rename("Kids_First_Biospecimen_ID_normal" = Kids_First_Biospecimen_ID) %>%
  left_join(hist[,c("Kids_First_Biospecimen_ID_normal", "plot_group")], by = "Kids_First_Biospecimen_ID_normal") %>%
  count(plot_group, Hugo_Symbol) %>%
  left_join(hist_cts, by = "plot_group") %>%
  mutate(n_without = hist_n - n) %>%
  dplyr::rename("Hugo_symbol" = Hugo_Symbol) %>%
  left_join(pmbb_cts[,c("Hugo_symbol", "count_with_plp_PMBB_noTumor", "count_without_plp_PMBB_noTumor")], by = "Hugo_symbol") %>%
  distinct(plot_group, Hugo_symbol, .keep_all = TRUE) %>%
  mutate(oddsRatio = 0, pvalue = 0, CI_lower = 0, CI_upper = 0)

# run Fisher's exact test by cpg and plot group 

for (i in 1:nrow(hist_cpg_enr_pmbb)){
  if(!is.na(hist_cpg_enr_pmbb$count_with_plp_PMBB_noTumor[i])) {
    mat <- matrix(c(hist_cpg_enr_pmbb$n[i], hist_cpg_enr_pmbb$n_without[i],
                    hist_cpg_enr_pmbb$count_with_plp_PMBB_noTumor[i], hist_cpg_enr_pmbb$count_without_plp_PMBB_noTumor[i]),
                  2,2)
    ftest <- fisher.test(mat)
    hist_cpg_enr_pmbb$oddsRatio[i] <- ftest$estimate
    hist_cpg_enr_pmbb$pvalue[i] <- ftest$p.value
    hist_cpg_enr_pmbb$CI_lower[i] <- ftest$conf.int[1]
    hist_cpg_enr_pmbb$CI_upper[i] <- ftest$conf.int[2]
  }
}

# calculate FDR

hist_cpg_enr_pmbb <- hist_cpg_enr_pmbb %>%
  mutate(adjusted_p = p.adjust(hist_cpg_enr_pmbb$pvalue, method = "bonferroni"))

# save output 

write_tsv(hist_cpg_enr_pmbb, 
          file.path(results_dir, "PBTA-cpg-plp-enrichment-by-plotGroup-PMBB.tsv"))
```

Combine gene and gene-by-histology PLP enrichment results across control cohorts

```{r}

cpg_enr_all <- cpg_enr_gnomad %>%
  rename("count_with_plp_PBTA" = n,
         "count_without_plp_PBTA" = n_without,
         "oddsRatio_gnomad" = oddsRatio,
         "pvalue_gnomad" = pvalue,
         "CI_lower_gnomad" = CI_lower,
         "CI_upper_gnomad" = CI_upper,
         "adjusted_p_gnomad" = adjusted_p) %>%
  left_join(cpg_enr_pmbb[,c("Hugo_symbol", "count_with_plp_PMBB_noTumor",
                            "count_without_plp_PMBB_noTumor", "oddsRatio",
                            "pvalue", "CI_lower", "CI_upper", "adjusted_p")],
            by = "Hugo_symbol") %>%
  rename("oddsRatio_pmbb" = oddsRatio,
         "pvalue_pmbb" = pvalue,
         "CI_lower_pmbb" = CI_lower,
         "CI_upper_pmbb" = CI_upper,
         "adjusted_p_pmbb" = adjusted_p)

write_tsv(cpg_enr_all,
          file.path(results_dir, "cpg-plp-enrichment-merged.tsv"))


hist_cpg_enr_all <- hist_cpg_enr_gnomad %>%
  rename("count_with_plp_PBTA" = n,
         "count_without_plp_PBTA" = n_without,
         "plot_group_n" = hist_n,
         "oddsRatio_gnomad" = oddsRatio,
         "pvalue_gnomad" = pvalue,
         "CI_lower_gnomad" = CI_lower,
         "CI_upper_gnomad" = CI_upper,
         "adjusted_p_gnomad" = adjusted_p) %>%
  left_join(hist_cpg_enr_pmbb[,c("plot_group", "Hugo_symbol", "count_with_plp_PMBB_noTumor",
                            "count_without_plp_PMBB_noTumor", "oddsRatio",
                            "pvalue", "CI_lower", "CI_upper", "adjusted_p")],
            by = c("plot_group", "Hugo_symbol")) %>%
  rename("oddsRatio_pmbb" = oddsRatio,
         "pvalue_pmbb" = pvalue,
         "CI_lower_pmbb" = CI_lower,
         "CI_upper_pmbb" = CI_upper,
         "adjusted_p_pmbb" = adjusted_p)

write_tsv(hist_cpg_enr_all,
          file.path(results_dir, "plot-group-cpg-plp-enrichment-merged.tsv"))

```
