---
title: 'Mutational Signatures Analysis in HGG tumors'
output: 
  html_document:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
date: "2023"
---

Load packages and set directories

```{r load libraries}
suppressPackageStartupMessages({
  library(data.table)
  library(tidyverse)
  library(ggplot2)
  library(ggsci)
  library(ggpubr)
})

root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "dna-repair-variant-summary")
input_dir <- c(file.path(analysis_dir, "input"))

source(file.path(analysis_dir, "util", "mutsigs_functions.R"))
source(file.path(root_dir, "figures", "theme.R"))

```

Set paths for: 
- histologies file, 
- germline plp variant file,
- COSMICv3 mutational signature weights file,
- TMB file

```{r set paths}
histologies_file <- file.path(root_dir, "analyses", "collapse-tumor-histologies",
                              "results","germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

tmb_file <- file.path(data_dir, "snv-mutation-tmb-coding.tsv")
```

Read in histologies, germline PLP, and tmb file. Subset histologies file to include only HGG tumors

```{r read files}
hist <- read_tsv(histologies_file,show_col_types = FALSE) 

tmb <- read_tsv(tmb_file,show_col_types = FALSE) %>%
  dplyr::rename("Kids_First_Biospecimen_ID" = "Tumor_Sample_Barcode")
```


Comparison of mutational signatures between HGG samples with and without germline variants in DNA repair genes will be performed through multiple loops to change the following parameters: 
- DNA repair gene list: run comparative analyses using two separate DNA repair gene lists (one obtained from [Knijnenburg et al. 2021 paper](https://pubmed.ncbi.nlm.nih.gov/29617664/), another from [Broad GSEA gene lists](https://www.gsea-msigdb.org/gsea/msigdb/human/genesets.jsp?collection=CP:KEGG))
- Pairwise comparisons; run separate comparative analyses of mutational signatures, considering 1) All patients with germline DNA repair variants vs. those without, 2) All patients with germline MMR gene variants vs. those without, 3) All patients with germline BRCA/BRCA-interacting gene variants vs. those without, and 4) all patients with other DNA repair germline variants vs. those without DNA repair variants. 

Consistent with [current practices](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6357558/pdf/bbx082.pdf), we set signature exposures < 0.06 to zero to exclude spurious signatures. 

```{r read mut sigs}

# Define cosmicV3 results files, one excluding and one including SBS39 from analysis 
cosmic_file <- file.path(input_dir, "COSMICv3.3_signature_exposures_SBS39_included.tsv")

# Read in COSMICv3 signature weights and set any exposures < 0.06 as 0
cosmicv3_sigs <- read_tsv(cosmic_file) %>%
tidyr::gather(-Kids_First_Biospecimen_ID, key = "signature", value = "exposure") %>%
mutate(exposure = case_when(
  exposure < 0.06 ~ 0, 
  TRUE ~ exposure
))
  
signatures <- cosmicv3_sigs %>%
  filter(exposure > 0) %>%
  distinct(signature) %>%
  pull(signature)
  
# Define MMR-associated and other DNA repair signatures
mmr_sigs <- c('SBS6', 'SBS14', 'SBS15', 'SBS20', 'SBS21', 'SBS26', 'SBS44')
repair_sigs <- c('SBS30', 'SBS36', 'SBS3', 'SBS9', 'SBS10a', 'SBS10b', 'SBS10c', 'SBS10d')

groups <- c("HGG, H3 wildtype", "DIPG or DMG",
            "Non-neoplastic tumor", "Mesenchymal tumor")
  
# Loop through two DNA repair gene lists
for (group in groups){
  
  if (group == "HGG, H3 wildtype"){
  
    sub_hist <- hist %>%
      filter(grepl("Other high-grade glioma", plot_group),
             grepl("wildtype", molecular_subtype))
    
  } else {
    
    sub_hist <- hist %>%
      dplyr::filter(plot_group == group)
    
  }
  
  # define gene list-specific results and plot directories
  results_dir <- file.path(analysis_dir, "results", group)
  plot_dir <- file.path(analysis_dir, "plots", group)
  
  if (!dir.exists(results_dir)){
    
    dir.create(results_dir)
    
  }
  
  if (!dir.exists(plot_dir)){
    
    dir.create(plot_dir)
    
  }
  
  dna_repair_plp_file <- file.path(analysis_dir, "results", "Knijnenburg_repair_genes", "dna_repair_germline_variant_plp_Knijnenburg_paper.tsv")
  dna_repair_sv_plp_file <- file.path(analysis_dir, "results", "Knijnenburg_repair_genes", "dna_repair_germline_sv_plp_Knijnenburg_paper.tsv")
  
  dna_repair_plp <- read_tsv(dna_repair_plp_file, show_col_types = FALSE)
  dna_repair_sv_plp <- read_tsv(dna_repair_sv_plp_file, show_col_types = FALSE)
  
  dna_repair_plp <- dna_repair_plp %>%
    dplyr::select(Kids_First_Biospecimen_ID_normal, gene_symbol_vep, 
                  BER, HR, MMR, NER, NHEJ) %>%
    bind_rows(dna_repair_sv_plp %>% dplyr::select(Kids_First_Biospecimen_ID_normal, gene_symbol_vep, 
                                                  BER, HR, MMR, NER, NHEJ))
  
  # obtain normal BS_ids for patients with germline PLP variant in: 1) any DNA repair gene, 2) MMR genes, 3) BRCA and BRCA_interacting genes, and 4) Other DNA repair genes, excluding MMR
  # Define BS_ids for normal and corresponding tumor IDs for each sample group 
  mmr_germline <- dna_repair_plp %>%
    filter(MMR == "Yes") %>%
    distinct(Kids_First_Biospecimen_ID_normal) %>%
    pull(Kids_First_Biospecimen_ID_normal)

  brca_germline <- dna_repair_plp %>%
    filter(grepl("BRCA|BARD1|BRIP1|ATM|CHEK2|PALB2|RAD51|MRE11A", gene_symbol_vep)) %>%
    distinct(Kids_First_Biospecimen_ID_normal) %>%
    pull(Kids_First_Biospecimen_ID_normal)
  
  ber_germline <- dna_repair_plp %>%
    filter(BER == "Yes") %>%
    distinct(Kids_First_Biospecimen_ID_normal) %>%
    pull(Kids_First_Biospecimen_ID_normal)

  # define list of BS_ids in each group, and names for subsequent plots
  sample_list <- list(mmr_germline, brca_germline, ber_germline)
  names(sample_list) <- c("MMR", "BRCA", "BER")
  
  plot_names <- c("MMR", "HR", "BER")
  
  # Loop through sample lists to compare mutational signatures between group of interest and samples without DNA repair germline variants
  for (k in 1:length(sample_list)){
  
    repair_germline <- sample_list[[k]]
    
    # Define control BS_ids list
    ctrl_germline <- sub_hist %>%
      filter(!Kids_First_Biospecimen_ID_normal %in% repair_germline) %>%
      pull(Kids_First_Biospecimen_ID_normal)

    # obtain corresponding tumors IDs for `repair_germline` and `ctrl_germline` samples 
    repair_ids <- sub_hist %>%
    filter(Kids_First_Biospecimen_ID_normal %in% repair_germline) %>%
    pull(Kids_First_Biospecimen_ID_tumor)
    
    ctrl_ids <- sub_hist %>%
    filter(Kids_First_Biospecimen_ID_normal %in% ctrl_germline) %>%
    pull(Kids_First_Biospecimen_ID_tumor)
    
    # define data frame to stores mean mutational signature exposures, differences between groups, and corresponding test statistic p-values 
    cosmicv3_res <- data.frame(signature = signatures,
                             repair_exposure = rep(0, length(signatures)),
                             ctrl_exposure = rep(0, length(signatures)),
                             exposure_diff = rep(0, length(signatures)),
                             wilcox_stat = rep(0, length(signatures)),
                             pvalue = rep(0, length(signatures)))
  
    # loop through exposures to obtain group exposure means and differences between groups
    for (l in 1:length(signatures)){
      repair_exposures <- cosmicv3_sigs[cosmicv3_sigs$signature == signatures[l] & cosmicv3_sigs$`Kids_First_Biospecimen_ID` %in% repair_ids,]$exposure
      ctrl_exposures <- cosmicv3_sigs[cosmicv3_sigs$signature == signatures[l] & cosmicv3_sigs$`Kids_First_Biospecimen_ID` %in% ctrl_ids,]$exposure
      cosmicv3_res$repair_exposure[l] <- mean(repair_exposures)
      cosmicv3_res$ctrl_exposure[l] <- mean(ctrl_exposures)
      cosmicv3_res$exposure_diff[l] <- mean(repair_exposures) - mean(ctrl_exposures)
      cosmicv3_res$wilcox_stat[l] <- ifelse(!is.nan(mean(repair_exposures) - mean(ctrl_exposures)),
                                       wilcox.test(repair_exposures, ctrl_exposures, alternative = 'two.sided')$statistic,
                                       NA)
      cosmicv3_res$pvalue[l] <- ifelse(!is.nan(mean(repair_exposures) - mean(ctrl_exposures)),
                                       wilcox.test(repair_exposures, ctrl_exposures, alternative = 'two.sided')$p.value,
                                       NA)
    }
    
    # write file to output 
    write_tsv(cosmicv3_res, file.path(results_dir, paste0("cosmicv3_sig_diff_", 
                                                          names(sample_list)[k], 
                                                          "_vs_other.tsv")))
  
    # create barplot of mutational signature differences between groups 
    cosmicv3_res <- cosmicv3_res %>%
      filter(!is.na(pvalue)) %>%
      mutate(association = case_when(
        signature %in% mmr_sigs ~ 'MMR',
        signature %in% repair_sigs ~ 'Other repair',
        TRUE ~ 'Non-Repair'
      )) %>%
      arrange(desc(exposure_diff))
    
    cosmicv3_res %>%
      mutate(signature = factor(signature)) %>% 
      mutate(signature = fct_relevel(signature, cosmicv3_res$signature)) %>%
      
      ggplot(aes(x = signature, y = exposure_diff, fill = association)) +
        geom_bar(stat="identity", color = "black") +
        theme_minimal() +
        theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
              plot.title = element_text(hjust = 0.5, size = 18)) + 
        labs(x = 'COSMIC v3.3 signature', y = 'exposure difference',
             fill = 'sig etiology', title = glue::glue("{group}, {plot_names[k]} P/LP vs. no P/LP")) +
        geom_hline(yintercept = 0) + 
        geom_text(aes(x = signature, y = exposure_diff * 1.1,
                  label = ifelse(pvalue < 0.05, "*", "")),
                  size = 8) +
        scale_fill_npg()
      
      ggsave(
        file.path(plot_dir, paste0("cosmicv3_signature_diff_", names(sample_list)[k], "_vs_other.tiff")), 
        width = 17, height = 10, units = "cm")

  }
  
}
```

Run mutational signature exposure comparisons between samples with and without DNA repair germline variants using both DNA repair gene list sources (Knijnenburg et al. and GSEA Broad)


Plot exposures of signatures of interest by group (MMR germline variant samples, BRCA/BRCA-interacting germline variant samples, other DNA repair germline variant samples, no DNA repair germline variant samples)

```{r}
# Define signatures of interest to plot
plot_sigs <- c(mmr_sigs, "SBS3", "SBS5", "SBS9", "SBS30")

# Define MMR-associated and other DNA repair-associated mutational signatures
# mmr_sigs <- c('SBS6', 'SBS14', 'SBS15', 'SBS20', 'SBS21', 'SBS26', 'SBS44')
# repair_sigs <- c('SBS30', 'SBS36', 'SBS3', 'SBS9', 'SBS10a', 'SBS10b', 'SBS10c', 'SBS10d')

# create plots for each DNA repair gene list source
for (group in groups){
  
  if (group == "HGG, H3 wildtype"){
  
    sub_hist <- hist %>%
      filter(grepl("Other high-grade glioma", plot_group),
             grepl("wildtype", molecular_subtype))
    
  } else {
    
    sub_hist <- hist %>%
      dplyr::filter(plot_group == group)
    
  }
  
    # define gene list-specific results and plot directories
  results_dir <- file.path(analysis_dir, "results", group)
  plot_dir <- file.path(analysis_dir, "plots", group)
  
  dna_repair_plp_file <- file.path(analysis_dir, "results", "Knijnenburg_repair_genes", "dna_repair_germline_variant_plp_Knijnenburg_paper.tsv")
  dna_repair_sv_plp_file <- file.path(analysis_dir, "results", "Knijnenburg_repair_genes", "dna_repair_germline_sv_plp_Knijnenburg_paper.tsv")
  
  dna_repair_plp <- read_tsv(dna_repair_plp_file, show_col_types = FALSE)
  dna_repair_sv_plp <- read_tsv(dna_repair_sv_plp_file, show_col_types = FALSE)
  
  dna_repair_plp <- dna_repair_plp %>%
    dplyr::select(Kids_First_Biospecimen_ID_normal, gene_symbol_vep, 
                  BER, HR, MMR, NER, NHEJ) %>%
    bind_rows(dna_repair_sv_plp %>% dplyr::select(Kids_First_Biospecimen_ID_normal, gene_symbol_vep, 
                                                  BER, HR, MMR, NER, NHEJ))
  

  # Define BS_ids for normal and corresponding tumor IDs for each sample group 
  mmr_germline <- dna_repair_plp %>%
    filter(MMR == "Yes") %>%
    distinct(Kids_First_Biospecimen_ID_normal) %>%
    pull(Kids_First_Biospecimen_ID_normal)

  brca_germline <- dna_repair_plp %>%
    filter(grepl("BRCA|BARD1|BRIP1|ATM|CHEK2|PALB2|RAD51|MRE11A", gene_symbol_vep)) %>%
    distinct(Kids_First_Biospecimen_ID_normal) %>%
    pull(Kids_First_Biospecimen_ID_normal)
  
  ber_germline <- dna_repair_plp %>%
    filter(BER == "Yes") %>%
    distinct(Kids_First_Biospecimen_ID_normal) %>%
    pull(Kids_First_Biospecimen_ID_normal)
    
  mmr_ids <- sub_hist %>%
    filter(Kids_First_Participant_ID %in% Kids_First_Participant_ID[Kids_First_Biospecimen_ID_normal %in% mmr_germline]) %>%
    pull(Kids_First_Biospecimen_ID_tumor)

  brca_ids <- sub_hist %>%
    filter(Kids_First_Participant_ID %in% Kids_First_Participant_ID[Kids_First_Biospecimen_ID_normal %in% brca_germline]) %>%
    pull(Kids_First_Biospecimen_ID_tumor)
  
  ber_ids <- sub_hist %>%
    filter(Kids_First_Participant_ID %in% Kids_First_Participant_ID[Kids_First_Biospecimen_ID_normal %in% ber_germline]) %>%
    pull(Kids_First_Biospecimen_ID_tumor)

  
  # Loop through mutational signatures of interest to generate exposure violin plots by group
  for (k in 1:length(plot_sigs)){
    
    if (plot_sigs[k] == "SBS3"){
      
    sbs_df <- cosmicv3_sigs %>%
      filter(signature == plot_sigs[k] & Kids_First_Biospecimen_ID %in% sub_hist$Kids_First_Biospecimen_ID_tumor) %>%
      mutate(germline_variant = case_when(
        Kids_First_Biospecimen_ID %in% brca_ids ~ "HR",
        TRUE ~ "Non-HR"
      )) %>%
      mutate(germline_variant = factor(germline_variant)) %>% 
      mutate(germline_variant = fct_relevel(germline_variant,
                                            c("HR", "Non-HR"))
             )
    
    plot_exposure_violin(sbs_df, "germline_variant", "exposure", plot_sigs[k],
                         group = group)
        
      ggsave(
        file.path(plot_dir, paste0(plot_sigs[k], "_signature_byGroup_violinPlot.tiff")), 
        width = 3, height = 3.5)
      
    } else if (plot_sigs[k] == "SBS30"){
      
    sbs_df <- cosmicv3_sigs %>%
      filter(signature == plot_sigs[k] & Kids_First_Biospecimen_ID %in% sub_hist$Kids_First_Biospecimen_ID_tumor) %>%
      mutate(germline_variant = case_when(
        Kids_First_Biospecimen_ID %in% ber_ids ~ "BER",
        TRUE ~ "Non-BER"
      )) %>%
      mutate(germline_variant = factor(germline_variant)) %>% 
      mutate(germline_variant = fct_relevel(germline_variant,
                                            c("BER", "Non-BER"))
             )
    
    plot_exposure_violin(sbs_df, "germline_variant", "exposure", plot_sigs[k],
                         group = group)
        
      ggsave(
        file.path(plot_dir, paste0(plot_sigs[k], "_signature_byGroup_violinPlot.tiff")), 
        width = 3, height = 3.5)
      
    } else {
    
    # Create data frame for plotting
    sbs_df <- cosmicv3_sigs %>%
      filter(signature == plot_sigs[k] & Kids_First_Biospecimen_ID %in% sub_hist$Kids_First_Biospecimen_ID_tumor) %>%
      mutate(germline_variant = case_when(
        Kids_First_Biospecimen_ID %in% mmr_ids ~ "MMR",
        TRUE ~ "Non-MMR"
      )) %>%
      mutate(germline_variant = factor(germline_variant)) %>% 
      mutate(germline_variant = fct_relevel(germline_variant,
                                            c("MMR", "Non-MMR"))
             )
    
    if (sum(sbs_df$germline_variant == "MMR") > 0 ){
    
      plot_exposure_violin(sbs_df, "germline_variant", "exposure", plot_sigs[k],
                           group = group)
          
      ggsave(
        file.path(plot_dir, paste0(plot_sigs[k], "_signature_byGroup_violinPlot.tiff")), 
        width = 3, height = 3.5)
        
    }
    
    }
    
  }
  

  # Create table of summed MMR deficiency-associated signatures across all samples
  mmr_df <- cosmicv3_sigs %>%
      filter(signature %in% mmr_sigs & Kids_First_Biospecimen_ID %in% sub_hist$Kids_First_Biospecimen_ID_tumor) %>%
      group_by(Kids_First_Biospecimen_ID) %>%
      summarize(sum = sum(exposure)) %>%
      mutate(germline_variant = case_when(
        Kids_First_Biospecimen_ID %in% mmr_ids ~ "MMR",
        TRUE ~ "Non-MMR"
      )) %>%
      mutate(germline_variant = factor(germline_variant)) %>% 
      mutate(germline_variant = fct_relevel(germline_variant,
                                            c("MMR", "Non-MMR"))
             )
  
  if (sum(mmr_df$germline_variant == "MMR") > 0 ){
    
    plot_exposure_violin(mmr_df, "germline_variant", "sum", "MMR",
                         group = group)
    
    ggsave(
      file.path(plot_dir, paste0("mmr_signature_byGroup_violinPlot.tiff")), 
      width = 3, height = 3.5)
    
    comparisons_mmr <- list(c("MMR", "Non-MMR"))
    
    tmb %>%
      dplyr::filter(Kids_First_Biospecimen_ID %in% sub_hist$Kids_First_Biospecimen_ID_tumor) %>%
      mutate(germline_variant = case_when(
          Kids_First_Biospecimen_ID %in% mmr_ids ~ "MMR",
          TRUE ~ "Non-MMR"
        )) %>%
      mutate(germline_variant = factor(germline_variant)) %>% 
      mutate(germline_variant = fct_relevel(germline_variant,
                                            c("MMR", "Non-MMR"))
             ) %>%
      
      ggplot(aes(x = germline_variant, y = log10(tmb), fill = germline_variant)) +
      geom_violin(binaxis = "y", stackdir = "center", 
                  show.legend = FALSE) +
      geom_boxplot(width=0.1, show.legend = FALSE) + 
      stat_compare_means(method = "wilcox",
                         comparisons = comparisons_mmr, size = 3,
                         method.args = list(alternative = "greater")) +
      labs(x = 'Germline Variant Class', y = "log10(TMB)",
           title = group) +
      theme_minimal() +
      theme(legend.position = 'none',
            text = element_text(size = 12)) +
      scale_x_discrete(labels = c(paste0("MMR P/LP\n (n=", length(mmr_ids), ")"), 
                                  paste0("MMR WT\n (n=", nrow(sub_hist) - length(mmr_ids), ")"))) +
      scale_fill_npg() +
      scale_y_continuous(expand = expansion(mult = .2)) +
      theme_Publication() +
      theme(plot.title = element_text(size = 15, hjust = 0.5),
            axis.title.x = element_text(size = 13)) 
          
    ggsave(
      file.path(plot_dir, paste0("tmb_byGroup_violinPlot.tiff")), 
      width = 3, height = 3.5)
    
  }
  
}
```

```{r}
sessionInfo()
```

