---
title: "Assess cancer predisposition gene expression in PBTA germline cohort"
output: 
  html_notebook:
  toc: TRUE
toc_float: TRUE
author: Ryan Corbett
---

This script compares genes expression z-scores in patients with vs. without germline pathogenic/likely pathogenic (P/LP) variants in PBTA germline cohort. 

# Load libraries
```{r load libraries, set up directories}
suppressPackageStartupMessages({
  library(ggplot2)
  library(tidyverse)
  library(ggpubr)
})

# Set up directories
root_dir <- rprojroot::find_root(rprojroot::has_dir(".git"))

data_dir <- file.path(root_dir, "data")
analysis_dir <- file.path(root_dir, "analyses", "two-hits")
input_dir <- file.path(analysis_dir, "input")
results_dir <- file.path(analysis_dir, "results")
plot_dir <- file.path(analysis_dir, "plots")

# If the directory does not exist, create it
if (!dir.exists(results_dir)) {
  dir.create(results_dir, recursive = TRUE)
}

# source publication theme
source(file.path(root_dir, "figures", "theme.R"))

# set seed to reproduce plots
set.seed(2023)

```

Set input file paths
```{r}

plp_file <- file.path(root_dir, "analyses", 
                      "variant-distribution", "input", 
                      "pbta_germline_plp_calls_autogvp_abridged_gnomad_popmax.tsv")

plp_sv_file <- file.path(root_dir, "analyses", 
                         "germline-sv", "input", 
                         "pbta_germline_svs.tsv")

histologies_file <- file.path(root_dir, "analyses",
                           "collapse-tumor-histologies",
                           "input", "histologies.tsv")

hist_cohort_file <- file.path(root_dir, "analyses", 
                              "collapse-tumor-histologies", "results", 
                              "germline-primary-plus-tumor-histologies-plot-groups-clin-meta.tsv")

expr_file <- file.path(data_dir, "gene-expression-rsem-tpm-collapsed.rds")

# cpg file
cpg_file <- file.path(root_dir, "analyses", "oncokb-annotation", "input", "cpg.txt")
hgat_file <- file.path(root_dir, "analyses", "variant-distribution", "input", "hgat_goi_list.tsv")
lgat_file <- file.path(root_dir, "analyses", "variant-distribution", "input", "lgat_goi_list.tsv")
embryonal_file <- file.path(root_dir, "analyses", "variant-distribution", "input", "embryonal-tumor_goi_list.tsv")
other_file <- file.path(root_dir, "analyses", "variant-distribution", "input", "other_goi_list.tsv")

```

# Load files
```{r load files}

hist <- read_tsv(hist_cohort_file)

# CPGs
cpgs <- read_lines(cpg_file)
hgat_drivers <- read_lines(hgat_file)
lgat_drivers <- read_lines(lgat_file)
embryonal_drivers <- read_lines(embryonal_file)
other_drivers <- read_lines(other_file)
somatic_drivers <- c(hgat_drivers, lgat_drivers, embryonal_drivers, other_drivers)

```

Read in germline plp file, and append coordinate and ref/alt allele info
```{r format germline data}

# Load abridged P-LP file and append variant coordinates
germline <- read_tsv(plp_file) %>%
  dplyr::rename(Hugo_Symbol = gene_symbol_vep) 

germline_sv <- read_tsv(plp_sv_file) %>%
  dplyr::rename(Hugo_Symbol = Hugo_Symbol_cpg)

```

Extract tumor RNA IDs from hist file, and append to `germline`
```{r get matched tumor RNA IDs}

rna_specimens <- read_tsv(file.path(data_dir, "independent-specimens.rnaseqpanel.primary-plus.tsv")) %>%
  dplyr::rename(Kids_First_Biospecimen_ID_rna = Kids_First_Biospecimen_ID)

hist <- hist %>%
  left_join(rna_specimens[,c("Kids_First_Participant_ID", "Kids_First_Biospecimen_ID_rna")])

# filter PLP calls for cpgs, and add plot_group and molecular_subtype
germline <- germline %>%
  filter(Hugo_Symbol %in% c(cpgs, somatic_drivers)) %>%
  left_join(hist[,c("Kids_First_Biospecimen_ID_normal", "Kids_First_Participant_ID", 
                    "Kids_First_Biospecimen_ID_rna", "plot_group", "molecular_subtype")]) %>%
  dplyr::select(Kids_First_Biospecimen_ID_normal, Kids_First_Participant_ID, 
                Kids_First_Biospecimen_ID_rna, 
                plot_group, molecular_subtype,
                Hugo_Symbol, autogvp_call, autogvp_call_reason,
                chr, start, ref, alt,
                HGVSc, HGVSp) %>%
  distinct(Kids_First_Biospecimen_ID_normal, Kids_First_Biospecimen_ID_rna, Hugo_Symbol, start, .keep_all = T)

germline_sv <- germline_sv %>%
  left_join(hist[,c("Kids_First_Biospecimen_ID_normal", "Kids_First_Participant_ID", 
                    "Kids_First_Biospecimen_ID_rna", "plot_group", "molecular_subtype")]) %>%
  dplyr::select(Kids_First_Biospecimen_ID_normal, Kids_First_Participant_ID, 
                Kids_First_Biospecimen_ID_rna, 
                plot_group, molecular_subtype,
                Hugo_Symbol,
                Chromosome, Start, End, Length,
                Type, Classification) %>%
  distinct(Kids_First_Biospecimen_ID_normal, Kids_First_Biospecimen_ID_rna, Hugo_Symbol, .keep_all = T)

```

Read in gene expression RDS file, and calculate z-scores
```{r format expression data}
expr <- readRDS(expr_file)
# subset `expr` to only include BS IDs in hist
subset_expr <- expr[,intersect(hist$Kids_First_Biospecimen_ID_rna, colnames(expr))]

# log2(x + 1) transform the expression matrix
log_expression <- log2(subset_expr + 1)

z_scored_expression <- log_expression %>%
  as.data.frame() %>%
  rownames_to_column("Hugo_Symbol") %>%
  gather(key = "Kids_First_Biospecimen_ID_rna", value = "expr", -Hugo_Symbol) %>%
  left_join(hist %>% dplyr::select(Kids_First_Biospecimen_ID_rna, Kids_First_Biospecimen_ID_normal,
                                   plot_group, plot_group_hex)) %>%
  group_by(Hugo_Symbol, plot_group) %>%
  dplyr::filter(Hugo_Symbol %in% germline$Hugo_Symbol) %>%
  dplyr::mutate(expr_zscore = scale(expr)) %>%
  dplyr::mutate(expr_zscore = as.numeric(expr_zscore))


write_tsv(z_scored_expression,
          file.path(results_dir, "pbta-germline-gene-expr-zscores.tsv"))
```

Append expression z-scores to genes with P/LP variants across cohort, and write to output
```{r get expression z-scores}

germline_somatic_expr <- germline %>%
  left_join(z_scored_expression)

germline_somatic_expr %>%
  dplyr::filter(Hugo_Symbol %in% cpgs) %>%
  dplyr::mutate(expr_zscore = as.numeric(expr_zscore)) %>%
  write_tsv(file.path(results_dir, "germline-somatic-expr-cpgs.tsv"))

germline_somatic_expr %>%
  dplyr::filter(!Hugo_Symbol %in% cpgs) %>%
  dplyr::mutate(expr_zscore = as.numeric(expr_zscore)) %>%
  write_tsv(file.path(results_dir, "germline-somatic-expr-somatic-non-cpgs.tsv"))

germline_sv_somatic_expr <- germline_sv %>%
  left_join(z_scored_expression) %>%
  dplyr::mutate(expr_zscore = as.numeric(expr_zscore)) %>%
  write_tsv(file.path(results_dir, "germline-sv-somatic-expr-cpgs.tsv"))
```


```{r}
expr_plot_df <- germline_somatic_expr %>%
  dplyr::filter(Hugo_Symbol %in% cpgs) %>%
  dplyr::select(Kids_First_Biospecimen_ID_rna, plot_group,
                Hugo_Symbol, expr_zscore) %>%
  bind_rows(germline_sv_somatic_expr %>%
              dplyr::select(Kids_First_Biospecimen_ID_rna, plot_group,
                            Hugo_Symbol, expr_zscore)) %>%
  left_join(hist %>% dplyr::select(plot_group, plot_group_hex)) %>%
  distinct() %>%
  dplyr::mutate(gene_text = case_when(
    expr_zscore > 2 | expr_zscore < -2 ~ Hugo_Symbol,
    TRUE ~ ""
  ))

plot_group_palette <- expr_plot_df$plot_group_hex
names(plot_group_palette) <- expr_plot_df$plot_group

expr_plot_df %>%
  dplyr::mutate(plot_group = fct_relevel(plot_group, 
                                         rev(sort(unique(plot_group))))) %>%
  
  ggplot(aes(x = expr_zscore, y = plot_group, label = gene_text)) + 
  geom_jitter(aes(fill = plot_group),
              shape = 21, width = 0.05,
              height = 0.2, 
              size = 2.5, alpha = 0.8,
              show.legend = FALSE) +
  geom_violin(aes(fill = plot_group),
              alpha = 0.2, outlier.shape = NA,
               show.legend = FALSE) +
  geom_label_repel(aes(label = gene_text),
                  size = 2.5,
                  box.padding   = 0.35, 
                  label.padding = 0.1,
                  point.padding = 0.25,
                  segment.color = 'grey50',
                  min.segment.length = 0.1) +
  geom_vline(xintercept = c(-2, 0, 2), linetype = "dashed", color = "gray60") + 
  annotate("text", x = 4, 
           y = min(expr_plot_df$plot_group), 
           label = glue::glue("N={sum(expr_plot_df$expr_zscore > 2 & !is.na(expr_plot_df$expr_zscore))}"), 
           color = "black", size = 5, hjust = 1, vjust = 1) +
    annotate("text", x = -3, 
           y = min(expr_plot_df$plot_group), 
           label = glue::glue("N={sum(expr_plot_df$expr_zscore < -2 & !is.na(expr_plot_df$expr_zscore))}"), 
           color = "black", size = 5, hjust = 1, vjust = 1) +
  scale_fill_manual(values = plot_group_palette, 
                     breaks = sort(names(plot_group_palette))) +
  labs(y = NULL, x = "TPM z-score") +
  xlim(c(-6,4)) +
  theme_Publication()

ggsave(file.path(plot_dir, "germline_plp_expr_zscores.pdf"),
       width = 6, height = 6.5)
```

For assessing differences in gene expression between patients with vs. without P/LP variants, we will take a single matched tumor RNA-seq sample for each patient from independent specimens file
```{r get mean expression by P/LP status}
germline_all <- germline %>%
  dplyr::select(Kids_First_Biospecimen_ID_normal, Hugo_Symbol, plot_group) %>%
  bind_rows(germline_sv %>% dplyr::select(Kids_First_Biospecimen_ID_normal, Hugo_Symbol, plot_group))

z_scored_expression <- z_scored_expression %>%
  # Add BS normal ids
  dplyr::mutate(sample_gene = glue::glue("{Kids_First_Biospecimen_ID_normal}:{Hugo_Symbol}")) %>%
  # add column indicating if patient has P/LP variant in gene
  dplyr::mutate(is_plp = case_when(
    sample_gene %in% glue::glue("{germline_all$Kids_First_Biospecimen_ID_normal}:{germline_all$Hugo_Symbol}") ~ "Yes",
    TRUE ~ "No"
  )) %>%
  distinct(Kids_First_Biospecimen_ID_rna, Hugo_Symbol,
           .keep_all = TRUE)
```

Calculate mean expression z-scores by gene and P/LP status, and identify genes with significant differences in expression between groups
```{r}

# calculate mean z-scored expression by gene and sample
mean_expr <- z_scored_expression %>%
  group_by(Hugo_Symbol, is_plp) %>%
  summarise(mean_zscore_expr = mean(expr_zscore))

# filter cpg list for those with P/LP variant in 4 or more patients
cpgs_test <- germline_all %>%
  dplyr::filter(Hugo_Symbol %in% cpgs) %>%
  distinct(Kids_First_Biospecimen_ID_normal, Hugo_Symbol, .keep_all = T) %>%
  count(Hugo_Symbol) %>%
  filter(n >= 4) %>%
  pull(Hugo_Symbol)

# create empty data frame to store expr diff results
expr_diff_df <- data.frame(Hugo_Symbol = cpgs_test,
                           mean_zscore_plp = rep(0, length(cpgs_test)),
                           mean_zscore_no_plp = rep(0, length(cpgs_test)),
                           pvalue = rep(0, length(cpgs_test)))

# loop through genes and assess significance of expr diff between pts with vs. without P/LP variants
for (i in 1:length(cpgs_test)) {
  
  expr_diff_df$mean_zscore_plp[i] <- mean_expr$mean_zscore_expr[mean_expr$Hugo_Symbol == cpgs_test[i] & mean_expr$is_plp == "Yes"]
  expr_diff_df$mean_zscore_no_plp[i] <- mean_expr$mean_zscore_expr[mean_expr$Hugo_Symbol == cpgs_test[i] & mean_expr$is_plp == "No"]
  expr_diff_df$pvalue[i] <- wilcox.test(z_scored_expression$expr_zscore[z_scored_expression$Hugo_Symbol == cpgs_test[i] & z_scored_expression$is_plp == "Yes"],
       z_scored_expression$expr_zscore[z_scored_expression$Hugo_Symbol == cpgs_test[i] & z_scored_expression$is_plp == "No"])$p.value
}

write_tsv(expr_diff_df, 
          file.path(results_dir, "gene-expr-diff-plp-vs-no-plp.tsv"))
```

Plot gene expression z-scores of selected CPGs by P/LP status
```{r plot expression by P/LP status}
z_scored_expression <- z_scored_expression %>%
  # Only samples with P/LP variants in genes will be colored by plot group
  dplyr::mutate(plp_group = case_when(
    is_plp == "No" ~ NA_character_,
    is_plp == "Yes" ~ plot_group
  )) %>%
  distinct(Kids_First_Biospecimen_ID_rna, Hugo_Symbol, .keep_all = TRUE)
```

Plot normalized expression within histologies, for genes that have P/LP variants in >= 3 patients within histology group
```{r plot expr by P/LP status within hist}
# define group_gene column
z_scored_expression <- z_scored_expression %>%
  dplyr::mutate(group_gene = glue::glue("{Hugo_Symbol}: {plot_group}"))

# Identify histology-gene pairs for which there are at least 3 patients with P/LP variants in group
plot_group_gene <- z_scored_expression %>%
  distinct(Kids_First_Biospecimen_ID_normal, Hugo_Symbol, .keep_all = TRUE) %>%
  filter(is_plp == "Yes") %>%
  count(group_gene) %>%
  dplyr::filter(n >= 2) %>%
  pull(group_gene)

compare_means_result <- z_scored_expression %>%
  dplyr::filter(group_gene %in% plot_group_gene) %>%
  group_by(group_gene) %>%
  summarize(p_value = wilcox.test(expr_zscore ~ is_plp)$p.value)

to_plot <- compare_means_result %>%
  dplyr::filter(p_value < 0.05) %>%
  pull(group_gene)

# plot expr z-scores by P/LP status within histologies for selected gene-histology pairs
z_scored_expression %>%
  dplyr::filter(group_gene %in% to_plot) %>%
  
   ggplot(aes(x = is_plp, y = expr_zscore)) + 
    geom_jitter(aes(fill = plot_group), shape = 21, width = 0.2, size = 2.5, alpha = 0.85) +
    geom_boxplot(alpha = 0.05, outlier.shape = NA) +
    stat_compare_means(method = "wilcox",
                       comparisons = list(c("Yes", "No")),
                       method.args = list(alternative = "two.sided")) + 
    labs(x = "Germline P/LP variant", y = "TPM z-score", fill = "Histology") +
    ylim(c(-5,5.75)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray") + 
    scale_fill_manual(values = plot_group_palette, 
                      breaks = sort(names(plot_group_palette))) +
    facet_wrap(~group_gene, nrow = 2,
               labeller = labeller(group_gene = label_wrap_gen(18))) + 
    theme_Publication()

  ggsave(file.path(plot_dir, "hist-gene-expr-plp-vs-noplp.pdf"),
                   width = 8, height = 4)
```

We can further separate samples based on whether matched tumors have detected oncogenic/likely oncogenic SNV, which may influence gene expression
```{r get ONCO SNV status}
# read in MAF file with oncokb annotation
maf <- read_tsv(file.path(root_dir, "analyses", "oncokb-annotation", "results", "snv-consensus-plus-hotspots-goi-oncokb.maf.tsv")) %>%
  filter(ONCOGENIC %in% c("Likely Oncogenic", "Oncogenic"))

loh <- read_tsv(file.path(results_dir, "germline-somatic-cnv-loh-cpgs.tsv")) %>%
  dplyr::filter(Kids_First_Biospecimen_ID_tumor %in% hist$Kids_First_Biospecimen_ID_tumor,
                loh_score >= 0.15)

second_hits_pathway <- read_tsv(file.path(results_dir, "germline_somatic_second_pathway_hits.tsv"))

# append matched tumor genomic ID to z_scored_expression
z_scored_expression <- z_scored_expression %>%
  left_join(hist[,c("Kids_First_Biospecimen_ID_normal", "Kids_First_Biospecimen_ID_tumor")])

z_scored_expression <- z_scored_expression %>%
  dplyr::mutate(second_hit = case_when(
    glue::glue("{Kids_First_Biospecimen_ID_tumor}:{Hugo_Symbol}") %in% c(glue::glue("{maf$Tumor_Sample_Barcode}:{maf$Hugo_Symbol}"),
                                                                         glue::glue("{loh$Kids_First_Biospecimen_ID_tumor}:{loh$Hugo_Symbol}"),
glue::glue("{second_hits_pathway$Kids_First_Biospecimen_ID_tumor_maf}:{second_hits_pathway$Hugo_Symbol}")) ~ "Yes",
    TRUE ~ "No"
  ))

z_scored_expression %>%
  dplyr::filter(is_plp == "Yes") %>%
  ggplot(aes(x = second_hit, y = expr_zscore)) +
  geom_jitter(aes(fill = plot_group), shape = 21, 
              width = 0.2, size = 2.5, alpha = 0.85,
              show.legend = FALSE) +
  geom_boxplot(alpha = 0.05, outlier.shape = NA) +
  stat_compare_means(method = "wilcox",
                     comparisons = list(c("Yes", "No")),
                       method.args = list(alternative = "less")) + 
  labs(x = "Somatic Second Hit", y = "TPM z-score", fill = "Histology") +
  ylim(c(-6,4)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") + 
  scale_fill_manual(values = plot_group_palette, 
                    breaks = sort(names(plot_group_palette))) +
  theme_Publication()

ggsave(file.path(plot_dir, "gene-expr-second-hit-vs-none.pdf"),
                   width = 3.5, height = 4)
```

Print session info
```{r session info}
sessionInfo()
```